### 2024-06-08 products/manuscript/project-proposal.qmd
Proposal Commit

---
title: "Life Expectancy"
subtitle: "PArt 1: Project Proposal"
author: Sean O'Sullivan
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

## Description of data and data source
The data set(s) that I'd liked to examine are a handful of annotated data sets pertaining to fraudulent (spam) emails. They come from a handful of sources, te largest of the bunch is from the Conference on Email and Anti-Spam. That data set and a pair of others found on [Kaggle](https://www.kaggle.com/datasets/naserabdullahalam/phishing-email-dataset/data?select=CEAS_08.csv) contain text information from the subjects and bodies of the emails as well as the senders, receivers, timestamps, and whether or not URLs are contained within.

## Questions/Hypotheses to be addressed
I'd like to explore the features that are prevalent in the identification of spam emails. Utilizing text analysis in addition to the other email attributes supplied my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for automated detection as used by some popular email providers and large internal IT departments across the corporate landscape.

I expect that a number of keywords common in spam emails will be relevant, but it's also likely that potential sender domains, whether the sender and receiver domain are the same, and what day of week or time of day the emails are sent are also highly correlated with intentionally spammy behavior.

While there may be some models better suited for this than others I intend to attempt a handful of classifiers and parameter hyper-tuning in order to compare and determine the best overall predictor of spam emails.

### 2024-06-08 products/manuscript/project-proposal.qmd
Proposal Commit

@@
 title: "Life Expectancy"
-subtitle: "PArt 1: Project Proposal"
+subtitle: "Part 1: Project Proposal"
 author: Sean O'Sullivan

### 2024-06-08 products/manuscript/project-proposal.qmd
Proposal Commit

@@
 ---
-title: "Life Expectancy"
+title: "Phishing for Spam"
 subtitle: "Part 1: Project Proposal"

### 2024-06-15 code/eda-code/eda.qmd
Loading, cleaning, and exploring the datasets.

---
title: "Exploratoray Analysis"
date: today
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(forcats)
library(plyr)
library(ggthemes)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
mydata$label = fct_recode(mydata$label, "No" = "0", "Yes" = "1")
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here::here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results for each of the plots to the `results` folder.

# Data exploration through figures

Density plots for the continuous outcomes -- email body length first.

```{r}
p1 <- mydata %>% ggplot(aes(x = bodylength, fill = label)) +
      geom_density(position = "dodge", alpha = 0.5) +
      xlim(0,7500) + xlab("Length") + ylab("Density") +
      labs(title = "Body Distribution", caption = "x-axis limited to a maximum of 7,500", fill = "Spam") +
      theme_tufte() +
      theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
plot(p1)
figure_file = here::here("results", "figures","bodylength-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now subject line length.

```{r}
p2 <- mydata %>% ggplot(aes(x=subjectlength, fill = label)) +
      geom_density(position = "dodge", alpha = 0.5) +
      xlim(0,150) + xlab("Length") + ylab("Density") +
      labs(title = "Subject Line Distribution", caption = "x-axis limited to a maximum of 150", fill = "Spam") +
      theme_tufte() +
      theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
plot(p2)
figure_file = here::here("results", "figures","subjectlength-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now the most common sender domains for spam emails.

```{r}
p3 <- mydata[which(mydata$label == "Yes" & !(mydata$senderdomain == "")),] %>%
      dplyr::count(senderdomain) %>%
      arrange(desc(n)) %>%
      head(25) %>%
      ggplot(aes(x = n, y = fct_reorder(senderdomain, n))) +
      geom_col(fill = "turquoise3", colour = "black", alpha = 0.65) +
      labs(title = "Spam Emails by Sender Domain") +
      ylab("") + xlab("Count") +
      theme_tufte() +
      theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.position = "none")
plot(p3)
figure_file = here::here("results", "figures", "spam-by-domain.png")
ggsave(filename = figure_file, plot=p3) 
```
```{r}
p3 <- mydata[which(mydata$label == "No" & !(mydata$senderdomain == "")),] %>%
      dplyr::count(senderdomain) %>%
      arrange(desc(n)) %>%
      head(25) %>%
      ggplot(aes(x = n, y = fct_reorder(senderdomain, n))) +
      geom_col(fill = "red3", colour = "black", alpha = 0.35) +
      labs(title = "Legit Emails by Sender Domain") +
      ylab("") + xlab("Count") +
      theme_tufte() +
      theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.position = "none")
plot(p3)
figure_file = here::here("results", "figures", "legit-by-domain.png")
ggsave(filename = figure_file, plot=p3) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

### 2024-06-15 products/manuscript/Manuscript.qmd
Loading, cleaning, and exploring the datasets.

---
title: "Fishing for Spam"
subtitle: "Part 1: Project Proposal"
author: Sean O'Sullivan
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

## General Background Information

*Provide enough background on your topic that others can understand the why and how of your analysis*

## Description of data and data source

The data set(s) that I'd liked to examine are a handful of annotated data sets pertaining to fraudulent (spam) emails. They come from a handful of sources, te largest of the bunch is from the Conference on Email and Anti-Spam. That data set and a pair of others found on [Kaggle](https://www.kaggle.com/datasets/naserabdullahalam/phishing-email-dataset/data?select=CEAS_08.csv) contain text information from the subjects and bodies of the emails as well as the senders, receivers, timestamps, and whether or not URLs are contained within.

## Questions/Hypotheses to be addressed

I'd like to explore the features that are prevalent in the identification of spam emails. Utilizing text analysis in addition to the other email attributes supplied my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for automated detection as used by some popular email providers and large internal IT departments across the corporate landscape.

I expect that a number of keywords common in spam emails will be relevant, but it's also likely that potential sender domains, whether the sender and receiver domain are the same, and what day of week or time of day the emails are sent are also highly correlated with intentionally spammy behavior.

While there may be some models better suited for this than others I intend to attempt a handful of classifiers and parameter hyper-tuning in order to compare and determine the best overall predictor of spam emails.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

## Data import and cleaning

To begin we start by downloading our data files of interest from [Kaggle](https://www.kaggle.com/datasets/naserabdullahalam/phishing-email-dataset/data?select=CEAS_08.csv). While there are a handful provided at the prior link we're mostly interested in the 3 CSV files that contain 6 features in addition to the spam annotation labels. Next we'll load the data into R with the `read_csv` and `here` functions. Once each of the 3 files is individually loaded into R we append the separate dataframes to one another by way of the `dplyr::bind_rows` function and start to summarize our cumulative data so that we can assess what data cleaning steps and feature generation is necessary.

Our first step in cleaning the data centers around data type conversions, namely changing the `urls` and `label` variables to be factors rather than character data types. Next, we proceeded with cleaning and extricating sender and receiver emails from their contact names where appropriate. In order to achieve this we utilized the `stringr` package and regular expressions to isolate the email addresses from the contact and mutate the original columns accordingly. We'll employee a similar technique shortly to isolate just the domain names for both sender and receiver emails.

In addition to cleaning some of the existing features we also sought to generate a few simple features that we believe may be of use in our analysis and modeling stages further down the line. Specifically, we created features that measure the character length of both the subject line and email bodies as well as a boolean features that assesses whether or not the sender and receiver email were the same.

By the end of our data loading and processing we arrived at a dataset that contains 7 character columns, 2 numeric columns, 2 factor columns, and a single boolean column -- taking care to ensure that our steps prevented generating any missing values where there hadn't previously been any. Ultimately this makes for a dataset with 48,295 rows across 11 features and our variable of interest. Complete steps and code can be located in the `.../code/processingprocessingcode.R` file found within the project.

## Exploratory/Descriptive analysis

Once the data had been loaded and cleaned we began to examine some of the relationships between anticipated predictor variables and our variable of interest.

![](../../results/figures/bodylength-distribution.png){.lightbox} First examining the distribution of email body length between spam and legitimate emails we observe a tendency for spam emails to be shorter in length. However, interestingly the distribution of spam emails is bi-modal -- with a second peak at around the 4,000 character count -- which may indicate that spam vendors are aware of the identification potential of email length and have begun targeting a longer length to avoid notice.

![](../../results/figures/subjectlength-distribution.png){.lightbox} Similarly, when we examine subject line length we observe the same tendency for spam emails to trend shorter. Subject line does not share the same distinct bi-modal shape for spam emails that the bodies do, but subject line length distribution between both spam and legitimate emails overlap in large part which may reduce it's efficacy for classification modeling.

::: {layout-ncol="2"}
![](../../results/figures/spam-by-domain.png)

![](../../results/figures/legit-by-domain.png)
:::

Finally, we examined the most popular sender domain names for vendors of both spam and legitimate emails. In doing so we see some clear indications of which domains are favorable to spam vendors and which are not. Notably Gmail fails to make the top 25 domains for spam vendors despite being the far-and-away most common choice for legitimate emails. Similarly, Yahoo and Hotmail lead the pack for spam and are not among common domains for legitimate emails. This bodes well for the classification power of sender domain when we enter the modeling phase and begin to incorporate bag-of-words and text processing.

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2024-06-15 code/processing-code/processingcode.R

@@ -5,116 +5,93 @@
 #and saves it as Rds file in the processed-data folder
 #
 # Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
+# This is done so one can pull in the chunks of code into Quarto documents
 # see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
 ###############################
 
 
 ## ---- packages --------
-#load needed packages. make sure they are installed.
-library(readxl) #for loading Excel files
-library(dplyr) #for data processing/cleaning
-library(tidyr) #for data processing/cleaning
-library(skimr) #for nice visualization of data 
-library(here) #to set paths
+# load needed packages. make sure they are installed.
+library(dplyr) # for data processing/cleaning
+library(tidyverse) # for data processing/cleaning
+library(skimr) # for nice visualization of data 
+library(here) # to set paths
+library(lubridate) # to convert datetimes
+library(stringr) # to perform regex extractions
 
 
 ## ---- loaddata --------
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw-data","exampledata.xlsx")
-#load data. 
-#note that for functions that come from specific packages (instead of base R)
-# I often specify both package and function like so
-#package::function() that's not required one could just call the function
-#specifying the package makes it clearer where the function "lives",
-#but it adds typing. You can do it either way.
-rawdata <- readxl::read_excel(data_location)
-# We might also want to load the codebook to look at it
-codebook <- readxl::read_excel(data_location, sheet ="Codebook")
-
+# path to data
+# note the use of the here() package and not absolute paths
+file1_location <- here::here("data","raw-data","CEAS_08.csv")
+file2_location <- here::here("data","raw-data","Nigerian_Fraud.csv")
+file3_location <- here::here("data","raw-data","SpamAssasin.csv")
+# load data.
+file1 <- read.csv(file1_location)
+file2 <- read.csv(file2_location)
+file3 <- read.csv(file3_location)
+# append dataframes
+rawdata <- bind_rows(list(file1, file2, file3))
 
 ## ---- exploredata --------
 #take a look at the data
 dplyr::glimpse(rawdata)
 #another way to look at the data
 summary(rawdata)
-#yet another way to get an idea of the data
-head(rawdata)
 #this is a nice way to look at data
 skimr::skim(rawdata)
-#look in the Codebook for a variable explanation
-print(codebook)
 
 
[…]

### 2024-06-15 code/eda-code/eda.qmd
Loading, cleaning, and exploring the datasets.

@@
 Density plots for the continuous outcomes -- email body length first.
+We'll utilize the `ggplot2` visualization package, which offers a `geom_density` shape that is well suited to visualizing the distributions of multiple groups in the same plot.
+In our case that will be both spam and legitimate emails.
 
@@
 
-Now subject line length.
+And we'll produce the same visualization for subject line length.
 
@@
 ```
-
-Now the most common sender domains for spam emails.
+Now the most common sender domains for spam emails. We'll continue to utilize `ggplot2`, this time with the `geom_col` shape in order to produce a bar chart of counts.
 
@@
 ```
+And similarly we'll produce a bar chart for legitimate emails.
+
 ```{r}

### 2024-06-15 products/manuscript/Manuscript.qmd
Loading, cleaning, and exploring the datasets.

@@
 Ultimately this makes for a dataset with 48,295 rows across 11 features and our variable of interest.
-Complete steps and code can be located in the `.../code/processingprocessingcode.R` file found within the project.
+Complete steps and code can be located in the `.../code/processing/processingcode.R` file found within the project.
 
@@
 This bodes well for the classification power of sender domain when we enter the modeling phase and begin to incorporate bag-of-words and text processing.
+
+A complete set of exploratory visualization steps utilizing `ggplot2` can be located in the `.../code/eda-code/eda.qmd` file found within the project.
 

### 2024-07-18 code/analysis-code/statistical-analysis.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

---
title: "Statistical Analysis"
format: html
---

# Prepping Train Data for Modeling

## Loading Data and Packages

Our first step is to import our processed data set and load in all necessary packages and functions for the analysis. Additionally we'll set the seed for the entirety of this document for the purposes of ensuring the exact values of this analysis are reproducible.

```{r}
# load needed packages. make sure they are installed.
pacman::p_load(here, #for data loading/saving
               tidyverse, # for data manipulation and plotting
               skimr, # for data exploration and summary
               ggthemes, # for plot theming
               tidytext, # for text analysis
               quanteda, # for text analysis
               caret, # for model training
               e1071, # for model training
               randomForest, # for model training
               irlba, # for SVD
               doSNOW, # for parallel processing
               doParallel, # for parallel processing
               parallel) # for parallel processing

# path to data
data_location <- here::here("data","processed-data","processeddata.rds")
# load data
mydata <- readRDS(data_location)

# set global seed for reproducibility of exact values
set.seed(490)
```

## Train/Test Splits

Next we'll focus on splitting the data into train and test splits. We'll opt to create a partition that leaves us 30% of the processed data set for testing the generalizability of our final model. We'll also want to ensure that class representations are maintained proportionate to the original data set.

```{r}
# create smaller partition for the sake of being able to run this shit on my PC
subsample <- createDataPartition(mydata$label,
                                 times = 1,
                                 p = .1,
                                 list = FALSE)

sub.data <- mydata[subsample,]

# create index of partition
inTrain <- createDataPartition(sub.data$label, 
                               times = 1, 
                               p = 0.7,
                               list = FALSE)

# subset training data
train <- sub.data[inTrain,]

# subset test data
test <- sub.data[-inTrain,]
```

The output of our class proportions look correctly maintained by the `createDataPartition()` function so we can proceed with preparing our training data for future models.

## Controls and Functions

```{r}
# create CV folds and model controls
folds <- createMultiFolds(train$label,
                          k = 10,
                          times = 3)
ctrl <- trainControl(method = "repeatedcv",
                     number = 10,
                     repeats = 3,
                     index = folds)
gridSVM <- expand.grid(C = seq(.1, 1, by = .1),
                       sigma = c(0.01, 0.1, 0.5, 1))

# function definitions
# calculate term frequency
term.freq <- function(row) {
  row / sum(row)
}

# calculate inverse document frequency
inverse.doc.freq <- function(col) {
  corpus.size <- length(col)
  doc.count <- length(which(col > 0))

  log10(corpus.size / doc.count)
}

# calculate tf-idf
tf.idf <- function(tf, idf) {
  tf * idf
}

# project singular value decomp onto test split
projSVD <- function(trainirlba, testtfidf) {
  sigma.inverse <- 1 / trainirlba$d
  u.transpose <- t(trainirlba$u)
  projection <- t(sigma.inverse * u.transpose %*% t(testtfidf))
  projection <- as.data.frame(projection)
  names(projection) <- paste0("X", seq(1,300))
  return(projection)
}

# tune Random Forest model
customRF <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "rf",
                 trControl = ctrl,
                 tuneLength = 7)
  stopCluster(clus)
  return(model)
}

# tune an SVM model
customSVM <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "svmRadial",
                 trControl = ctrl,
                 tuneLength = 7,
                 tuneGrid = gridSVM)
  stopCluster(clus)
  return(model)
}
```

To reduce repetitive model configurations and parameter tuning grids we'll establish some variables and functions for later use in the analysis.

## Body Tokens and SVD

```{r}
# generate tokens from email bodies
train.body.tokens <- tokens(train$body, what = "word",
                       remove_numbers = TRUE,
                       remove_separators = TRUE,
                       remove_punct = TRUE,
                       remove_symbols = TRUE,
                       split_hyphens = FALSE,
                       split_tags = TRUE,
                       remove_url = FALSE)
```

Our first step in structuring text data requires that we tokenize the email bodies.

```{r}
# lowercase tokens
train.body.tokens <- tokens_tolower(train.body.tokens)

# remove stopwords
train.body.tokens <- tokens_select(train.body.tokens,
                                   stopwords(),
                                   selection = "remove")

# stem words
train.body.tokens <- tokens_wordstem(train.body.tokens,
                                     language = "english")
```

After which we'll proceed with standardizing and consolidating the tokens. In our case we'll be lowercasing all tokens, removing English stopwords, and consolidating the semantics by stemming the remaining tokens.

```{r}
# create document-feature matrix
train.body.tokens.dfm <- dfm(train.body.tokens,
                             tolower = FALSE)

# convert dfm to matrix
train.body.tokens.matrix <- as.matrix(train.body.tokens.dfm)
```

After we've tokenized we're now able to restructure the data as a document-feature matrix in preparation for calculating term frequencies and inverse document frequencies.

```{r}
# create the term frequency
train.body.tokens.tf <- apply(train.body.tokens.matrix, 1, term.freq)

# create the inverse document frequency
train.body.tokens.idf <- apply(train.body.tokens.matrix, 2, inverse.doc.freq)

# create the tf-idf
train.body.tokens.tfidf <- apply(train.body.tokens.tf, 2, tf.idf, idf = train.body.tokens.idf)
```

We proceed by calculating TF-IDFs for the email body token data. This is valuable because it standardizes the word occurrences and further helps to reduce the influence of over represented words on the model. The TF values will accomplish the former while the IDF values will accomplish the latter by discounting the tokens found consistently throughout the corpus of emails. In the case of tokens found in every email throughout the corpus the end result is effectively complete removal from the model akin to our earlier removal of stop words.

```{r}
# transpose tf-idf and handle incomplete cases
train.body.tokens.tfidf <- t(train.body.tokens.tfidf)
incomplete.cases <- which(!complete.cases(train.body.tokens.tfidf))
train.body.tokens.tfidf[incomplete.cases,] <- rep(0.0, ncol(train.body.tokens.tfidf))
```

Having reinterpreted our tokens as TF-IDFs we can transpose the matrix back to it's original orientation and fill in zeroes for any tokens that generated NA values (as a result of no representation of the given tokens within the email).

```{r}
# run singular value decomposition for body tokens
train.body.irlba <- irlba(t(train.body.tokens.tfidf), nv = 300, maxit = 600)

saveRDS(train.body.irlba, file = here("results", "large-files", "train-body-irlba.rds"))
```

Finally, we're left with a matrix containing 40,399 features. While many classification models can ultimately accommodate such intense dimensionality, the computational resource requirements are quite high and would therefore make it unfeasible to run these models with the limited memory and processing power available on consumer-grade hardware. Additionally, the sparsity of our matrix (\>99.9%) suggests that many of the features will not be particularly predictive of either classification on their own.

As a result, we'll attempt to reduce the dimensionality of our feature set by employing the `irlba` package to perform singular value decomposition with a target feature limit of the 300 most informative predictor combinations.

## Subject Tokens and SVD

*This section features the same steps as above, but now for the email subject text. Justifications for each process will be the same as before.*

```{r}
# generate tokens from email subjects
train.sub.tokens <- tokens(train$subject, what = "word",
                       remove_numbers = TRUE,
                       remove_separators = TRUE,
                       remove_punct = TRUE,
                       remove_symbols = TRUE,
                       split_hyphens = FALSE,
                       split_tags = TRUE,
                       remove_url = FALSE)
```

```{r}
# lowercase tokens
train.sub.tokens <- tokens_tolower(train.sub.tokens)

# remove stopwords
train.sub.tokens <- tokens_select(train.sub.tokens,
                                   stopwords(),
                                   selection = "remove")

# stem words
train.sub.tokens <- tokens_wordstem(train.sub.tokens,
                                     language = "english")
```

```{r}
# create document-feature matrix
train.sub.tokens.dfm <- dfm(train.sub.tokens,
                             tolower = FALSE)
```

```{r}
# convert dfm to matrix
train.sub.tokens.matrix <- as.matrix(train.sub.tokens.dfm)
```

```{r}
# create the term frequency
train.sub.tokens.tf <- apply(train.sub.tokens.matrix, 1, term.freq)
```

```{r}
# create the inverse document frequency
train.sub.tokens.idf <- apply(train.sub.tokens.matrix, 2, inverse.doc.freq)
```

```{r}
# create the tf-idf
train.sub.tokens.tfidf <- apply(train.sub.tokens.tf, 2, tf.idf, idf = train.sub.tokens.idf)
```

```{r}
# transpose tf-idf and handle incomplete cases
train.sub.tokens.tfidf <- t(train.sub.tokens.tfidf)
incomplete.cases <- which(!complete.cases(train.sub.tokens.tfidf))
train.sub.tokens.tfidf[incomplete.cases,] <- rep(0.0, ncol(train.sub.tokens.tfidf))
```

```{r}
# create single value decomp for subject tokens
train.sub.irlba <- irlba(t(train.sub.tokens.tfidf), nv = 300, maxit = 600)

saveRDS(train.sub.irlba, file = here("results", "large-files", "train-sub-irlba.rds"))
```

## Final Data Frames for Modeling

```{r}
####### Body Token Only #######
# create revised training data frame with body token SVD
train.body.svd <- data.frame(label = train$label,
                             train.body.irlba$v)
saveRDS(train.body.svd, file = here("results", "large-files", "train-body-svd.rds"))

# create revised training data frame with body token SVD and email length
train.body.svd.length <- data.frame(label = train$label,
                               length = train$bodylength,
                               train.body.irlba$v)
saveRDS(train.body.svd.length, file = here("results", "large-files", "train-body-svd-length.rds"))

# create revised training data frame 
# with body token SVD, email length, and sender domain
train.body.svd.length.sender <- data.frame(label = train$label,
                                           length = train$bodylength,
                                           senderdomain = train$senderdomain,
                                           train.body.irlba$v)
saveRDS(train.body.svd.length.sender, file = here("results", "large-files", "train-body-svd-length-sender.rds"))

####### Subject Token Only #######
# create revised training data frame with subject token SVD
train.sub.svd <- data.frame(label = train$label,
                            train.sub.irlba$v)
saveRDS(train.sub.svd, file = here("results", "large-files", "train-sub-svd.rds"))

# create revised training data frame with subject token SVD and email length
train.sub.svd.length <- data.frame(label = train$label,
                                   length = train$bodylength,
                                   train.sub.irlba$v)
saveRDS(train.sub.svd.length, file = here("results", "large-files", "train-sub-svd-length.rds"))

# create revised training data frame 
# with subject token SVD, email length, and sender domain
train.sub.svd.length.sender <- data.frame(label = train$label,
                                          length = train$bodylength,
                                          senderdomain = train$senderdomain,
                                          train.sub.irlba$v)
saveRDS(train.sub.svd.length.sender, file = here("results", "large-files", "train-sub-svd-length-sender.rds"))

####### All Tokens #######
# create revised training data frame with all tokens SVD
train.all.svd <- data.frame(label = train$label,
                            train.body.irlba$v,
                            train.sub.irlba$v)
saveRDS(train.all.svd, file = here("results", "large-files", "train-all-svd.rds"))

# create revised training data frame with all tokens SVD and email length
train.all.svd.length <- data.frame(label = train$label,
                                   length = train$bodylength,
                                   train.body.irlba$v,
                                   train.sub.irlba$v)
saveRDS(train.all.svd.length, file = here("results", "large-files", "train-all-svd-length.rds"))

# create revised training data frame 
# with all tokens SVD, email length, and sender domain
train.all.svd.length.sender <- data.frame(label = train$label,
                                          length = train$bodylength,
                                          senderdomain = train$senderdomain,
                                          train.body.irlba$v,
                                          train.sub.irlba$v)
saveRDS(train.all.svd.length.sender, file = here("results", "large-files", "train-all-svd-length-sender.rds"))
```

Having generated structured text data that is also manageable, we can now piece together different data frames consisting of the various feature combinations we would like to assess for modeling. Based on the relationships observed in our exploratory data analysis those features are the token data for both email subject and body, the body length of the email, and the sender domain for the email.

# Tune Models

## Random Forest Iterations

```{r}
###### Body Tokens Only ######
# tune a Random Forest model with body token SVD
rf.body.svd <- customRF(train.body.svd)

saveRDS(rf.body.svd, file = here("results", "large-files", "rf-body-svd.rds"))

# tune a Random Forest model with body token SVD and email length
rf.body.svd.length <- customRF(train.body.svd.length)

saveRDS(rf.body.svd.length, file = here("results", "large-files", "rf-body-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune a Random Forest model with body token SVD, email length, and sender domain
#rf.body.svd.length.sender <- customRF(train.body.svd.length.sender)

#saveRDS(rf.body.svd.length.sender, file = here("results", "large-files", "rf-body-svd-length-sender.rds"))

###### Subject Tokens Only ######
# tune a Random Forest model with subject token SVD
rf.sub.svd <- customRF(train.sub.svd)

saveRDS(rf.sub.svd, file = here("results", "large-files", "rf-sub-svd.rds"))

# tune a Random Forest model with subject token SVD and email length
rf.sub.svd.length <- customRF(train.sub.svd.length)

saveRDS(rf.sub.svd.length, file = here("results", "large-files", "rf-sub-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune a Random Forest model with subject token SVD, email length, and sender domain
#rf.sub.svd.length.sender <- customRF(train.sub.svd.length.sender)

#saveRDS(rf.sub.svd.length.sender, file = here("results", "large-files", "rf-sub-svd-length-sender.rds"))

###### All Tokens ######
# tune a Random Forest model with all token SVD
rf.all.svd <- customRF(train.all.svd)

saveRDS(rf.all.svd, file = here("results", "large-files", "rf-all-svd.rds"))

# tune a Random Forest model with all token SVD and email length
rf.all.svd.length <- customRF(train.all.svd.length)

saveRDS(rf.all.svd.length, file = here("results", "large-files", "rf-all-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune a Random Forest model with all token SVD, email length, and sender domain
#rf.all.svd.length.sender <- customRF(train.all.svd.length.sender)

#saveRDS(rf.all.svd.length.sender, file = here("results", "large-files", "rf-all-svd-length-sender.rds"))
```

## SVM Iterations

```{r}
###### Body Tokens Only ######
# tune an SVM model with body token SVD
svm.body.svd <- customSVM(train.body.svd)

saveRDS(svm.body.svd, file = here("results", "large-files", "svm-body-svd.rds"))

# tune an SVM model with body token SVD and email length
svm.body.svd.length <- customSVM(train.body.svd.length)

saveRDS(svm.body.svd.length, file = here("results", "large-files", "svm-body-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune an SVM model with body token SVD, email length, and sender domain
#svm.body.svd.length.sender <- customSVM(train.body.svd.length.sender)

#saveRDS(svm.body.svd.length.sender, file = here("results", "large-files", "svm-body-svd-length-sender.rds"))

###### Subject Tokens Only ######
# tune an SVM model with subject token SVD
svm.sub.svd <- customSVM(train.sub.svd)

saveRDS(svm.sub.svd, file = here("results", "large-files", "svm-sub-svd.rds"))

# tune an SVM model with subject token SVD and email length
svm.sub.svd.length <- customSVM(train.sub.svd.length)

saveRDS(svm.sub.svd.length, file = here("results", "large-files", "svm-sub-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune an SVM model with subject token SVD, email length, and sender domain
#svm.sub.svd.length.sender <- customSVM(train.sub.svd.length.sender)

#saveRDS(svm.sub.svd.length.sender, file = here("results", "large-files", "svm-sub-svd-length-sender.rds"))

###### All Tokens ######
# tune an SVM model with all token SVD
svm.all.svd <- customSVM(train.all.svd)

saveRDS(svm.all.svd, file = here("results", "large-files", "svm-all-svd.rds"))

# tune an SVM model with all token SVD and email length
svm.all.svd.length <- customSVM(train.all.svd.length)

saveRDS(svm.all.svd.length, file = here("results", "large-files", "svm-all-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune an SVM model with all token SVD, email length, and sender domain
#svm.all.svd.length.sender <- customSVM(train.all.svd.length.sender)

#saveRDS(svm.all.svd.length.sender, file = here("results", "large-files", "svm-all-svd-length-sender.rds"))
```

# Prepping Test Data for Predictions

## Body Tokens and SVD

```{r}
# generate tokens from test email bodies
test.body.tokens <- tokens(test$body, what = "word",
                       remove_numbers = TRUE,
                       remove_separators = TRUE,
                       remove_punct = TRUE,
                       remove_symbols = TRUE,
                       split_hyphens = FALSE,
                       split_tags = TRUE,
                       remove_url = FALSE)
```

```{r}
# lowercase tokens
test.body.tokens <- tokens_tolower(test.body.tokens)

# remove stopwords
test.body.tokens <- tokens_select(test.body.tokens,
                                   stopwords(),
                                   selection = "remove")

# stem words
test.body.tokens <- tokens_wordstem(test.body.tokens,
                                     language = "english")
```

```{r}
# create document-feature matrix
test.body.tokens.dfm <- dfm(test.body.tokens,
                             tolower = FALSE)
```

```{r}
# project the feature set from the train split onto the test split
test.body.tokens.dfm <- dfm_match(test.body.tokens.dfm, features = featnames(train.body.tokens.dfm))
```

```{r}
# convert dfm to matrix
test.body.tokens.matrix <- as.matrix(test.body.tokens.dfm)
```

```{r}
test.body.tokens.tf <- apply(test.body.tokens.matrix, 1, term.freq)
```

```{r}
test.body.tokens.tfidf <- apply(test.body.tokens.tf, 2, tf.idf, idf = train.body.tokens.idf)
```

```{r}
# transpose and fix missing values
test.body.tokens.tfidf <- t(test.body.tokens.tfidf)
test.body.tokens.tfidf[is.na(test.body.tokens.tfidf)] <- 0.0
```

```{r}
# project the SVD from the train split onto the test split
test.body.svd.proj <- projSVD(train.body.irlba, test.body.tokens.tfidf)
```

## Subject Tokens and SVD

```{r}
# generate tokens from test email bodies
test.sub.tokens <- tokens(test$subject, what = "word",
                          remove_numbers = TRUE,
                          remove_separators = TRUE,
                          remove_punct = TRUE,
                          remove_symbols = TRUE,
                          split_hyphens = FALSE,
                          split_tags = TRUE,
                          remove_url = FALSE)
```

```{r}
# lowercase tokens
test.sub.tokens <- tokens_tolower(test.sub.tokens)

# remove stopwords
test.sub.tokens <- tokens_select(test.sub.tokens,
                                 stopwords(),
                                 selection = "remove")

# stem words
test.sub.tokens <- tokens_wordstem(test.sub.tokens,
                                   language = "english")
```

```{r}
# create document-feature matrix
test.sub.tokens.dfm <- dfm(test.sub.tokens,
                           tolower = FALSE)
```

```{r}
# project the feature set from the train split onto the test split
test.sub.tokens.dfm <- dfm_match(test.sub.tokens.dfm, features = featnames(train.sub.tokens.dfm))
```

```{r}
# convert dfm to matrix
test.sub.tokens.matrix <- as.matrix(test.sub.tokens.dfm)
```

```{r}
test.sub.tokens.tf <- apply(test.sub.tokens.matrix, 1, term.freq)
```

```{r}
test.sub.tokens.tfidf <- apply(test.sub.tokens.tf, 2, tf.idf, idf = train.sub.tokens.idf)
```

```{r}
# transpose and fix missing values
test.sub.tokens.tfidf <- t(test.sub.tokens.tfidf)
test.sub.tokens.tfidf[is.na(test.sub.tokens.tfidf)] <- 0.0
```

```{r}
# project the SVD from the train split onto the test split
test.sub.svd.proj <- projSVD(train.sub.irlba, test.sub.tokens.tfidf)
```

## Final Data Frames for Predictions

```{r}
####### Body Token Only #######
# create data frame of body tokens and labels
test.body.svd <- cbind(label = test$label,
                       test.body.svd.proj)

saveRDS(test.body.svd, file = here("results", "large-files", "test-body-svd.rds"))

# create data frame of body tokens and labels
test.body.svd.length <- cbind(label = test$label,
                              length = test$bodylength,
                              test.body.svd.proj)

saveRDS(test.body.svd.length, file = here("results", "large-files", "test-body-svd-length.rds"))

# create data frame of body tokens and labels
test.body.svd.length.sender <- cbind(label = test$label,
                                     length = test$bodylength,
                                     senderdomain = test$senderdomain,
                                     test.body.svd.proj)

saveRDS(test.body.svd.length.sender, file = here("results", "large-files", "test-body-svd-length-sender.rds"))

####### Subject Token Only #######
# create data frame of subject tokens and labels
test.sub.svd <- cbind(label = test$label,
                      test.sub.svd.proj)

saveRDS(test.sub.svd, file = here("results", "large-files", "test-sub-svd.rds"))

# create data frame of subject tokens and labels
test.sub.svd.length <- cbind(label = test$label,
                             length = test$bodylength,
                             test.sub.svd.proj)

saveRDS(test.sub.svd.length, file = here("results", "large-files", "test-sub-svd-length.rds"))

# create data frame of subject tokens and labels
test.sub.svd.length.sender <- cbind(label = test$label,
                                    length = test$bodylength,
                                    senderdomain = test$senderdomain,
                                    test.sub.svd.proj)

saveRDS(test.sub.svd.length.sender, file = here("results", "large-files", "test-sub-svd-length-sender.rds"))

####### All Tokens #######
# create data frame of all tokens and labels
test.all.svd <- cbind(label = test$label,
                      test.body.svd.proj,
                      test.sub.svd.proj)

saveRDS(test.all.svd, file = here("results", "large-files", "test-all-svd.rds"))

# create data frame of all tokens and labels
test.all.svd.length <- cbind(label = test$label,
                             length = test$bodylength,
                             test.body.svd.proj,
                             test.sub.svd.proj)

saveRDS(test.all.svd.length, file = here("results", "large-files", "test-all-svd-length.rds"))

# create data frame of all tokens and labels
test.all.svd.length.sender <- cbind(label = test$label,
                                    length = test$bodylength,
                                    senderdomain = test$senderdomain,
                                    test.body.svd.proj,
                                    test.sub.svd.proj)

saveRDS(test.all.svd.length.sender, file = here("results", "large-files", "test-all-svd-length-sender.rds"))
```

# Generate Predictions

```{r}
# predict test data with model
#rf.body.preds <- predict(rf.body, test.body.svd)
#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2024-07-18 code/analysis-code/CopyOfstatistical-analysis.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

--- 2024-07-18 code/analysis-code/statistical-analysis.qmd
@@
                tidyverse, # for data manipulation and plotting
-               skimr, # for data exploration and summary
-               ggthemes, # for plot theming
                tidytext, # for text analysis
@@
                randomForest, # for model training
+               xgboost, # for model training
                irlba, # for SVD
-               doSNOW, # for parallel processing
                doParallel, # for parallel processing
@@
 
+However, before we can proceed with either step we first need to ensure that the size of our data is manageable for the scope of our machine.
+Modeling a very large corpus of unstructured text data incurs high computational costs due to the increasing dimensionality of the data as we begin to structure it.
+As a result, the amount of data that can be modeled will vary significantly based on the computational resources available to you.
+
+At this time we've made the decision to examine a 10% sample of our overall data set for the purpose of modeling on our machine, but the proportion parameter for the subsample partition can be tweaked according to your available resources.
+
 ```{r}
-# create smaller partition for the sake of being able to run this shit on my PC
+# create smaller partition for the sake of being able
+# to run on consumer grade hardware 
 subsample <- createDataPartition(mydata$label,
@@
 
-# create index of partition
+# create index of train partition
 inTrain <- createDataPartition(sub.data$label, 
@@
 
+To reduce repetitive model configurations and parameter tuning grids we'll establish some variables and functions for later use in the analysis.
+
 ```{r}
@@
                      repeats = 3,
-                     index = folds)
+                     index = folds,
+                     allowParallel = TRUE)
 gridSVM <- expand.grid(C = seq(.1, 1, by = .1),
                        sigma = c(0.01, 0.1, 0.5, 1))
+gridXG <- expand.grid(nrounds = c(100, 200),
+                      max_depth = c(3, 6),
+                      eta = c(0.1, 0.3),
+                      gamma = c(0, 0.1),
+                      colsample_bytree = c(0.8),
+                      min_child_weight = c(1, 3),
+                      subsample = c(0.8))
 
@@
 # project singular value decomp onto test split
-projSVD <- function(trainirlba, testtfidf) {
+projSVD <- function(type, trainirlba, testtfidf) {
+
   sigma.inverse <- 1 / trainirlba$d
@@
   projection <- as.data.frame(projection)
+  if (type == "bod") {
     names(projection) <- paste0("X", seq(1,300))
+  } else {
+    names(projection) <- paste0("Z", seq(1,300))
+  }
   return(projection)
@@
   cores <- detectCores()
-  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
+  clus <- makeCluster(cores[1] - 1) # determine the optimal number
                                                    # of cores for parallel processing
@@
                  trControl = ctrl,
-                 tuneLength = 7)
+                 tuneLength = 7,
+                 verbose = FALSE)
   stopCluster(clus)
@@
                  tuneLength = 7,
-                 tuneGrid = gridSVM)
+                 tuneGrid = gridSVM,
+                 verbose = FALSE)
+  stopCluster(clus)
+  return(model)
+}
+
+# tune an XGBoost model
+customXG <- function(df) {
+  cores <- detectCores()
+  clus <- makeCluster(cores[1] - 1) # determine the optimal number
+                                                   # of cores for parallel processing
+  registerDoParallel(clus)
+  model <- train(label ~ ., 
+                 data = df,
+                 method = "xgbTree",
+                 trControl = ctrl,
+                 tuneGrid = gridXG,
+                 verbose = FALSE)
   stopCluster(clus)
@@
 
-To reduce repetitive model configurations and parameter tuning grids we'll establish some variables and functions for later use in the analysis.
+## Body Tokens and SVD
 
-## Body Tokens and SVD
+Our first step in structuring text data requires that we tokenize the email bodies.
 
@@
 
-Our first step in structuring text data requires that we tokenize the email bodies.
+After which we'll proceed with standardizing and consolidating the tokens.
+In our case we'll be lowercasing all tokens, removing English stopwords, and consolidating the semantics by stemming the remaining tokens.
 
@@
 
-After which we'll proceed with standardizing and consolidating the tokens.
-In our case we'll be lowercasing all tokens, removing English stopwords, and consolidating the semantics by stemming the remaining tokens.
+After we've tokenized we're now able to restructure the data as a document-feature matrix in preparation for calculating term frequencies and inverse document frequencies.
 
@@
 
-After we've tokenized we're now able to restructure the data as a document-feature matrix in preparation for calculating term frequencies and inverse document frequencies.
+We proceed by calculating TF-IDFs for the email body token data.
+This is valuable because it standardizes the word occurrences and further helps to reduce the influence of over represented words on the model.
+The TF values will accomplish the former while the IDF values will accomplish the latter by discounting the tokens found consistently throughout the corpus of emails.
+In the case of tokens found in every email throughout the corpus the end result is effectively complete removal from the model akin to our earlier removal of stop words.
 
@@
 
-We proceed by calculating TF-IDFs for the email body token data.
-This is valuable because it standardizes the word occurrences and further helps to reduce the influence of over represented words on the model.
-The TF values will accomplish the former while the IDF values will accomplish the latter by discounting the tokens found consistently throughout the corpus of emails.
-In the case of tokens found in every email throughout the corpus the end result is effectively complete removal from the model akin to our earlier removal of stop words.
+Having reinterpreted our tokens as TF-IDFs we can transpose the matrix back to it's original orientation and fill in zeroes for any tokens that generated NA values (as a result of no representation of the given tokens within the email).
 
@@
 
-Having reinterpreted our tokens as TF-IDFs we can transpose the matrix back to it's original orientation and fill in zeroes for any tokens that generated NA values (as a result of no representation of the given tokens within the email).
+Finally, we're left with a matrix containing 40,399 features.
+While many classification models can ultimately accommodate such intense dimensionality, the computational resource requirements are quite high and would therefore make it unfeasible to run these models with the limited memory and processing power available on consumer-grade hardware.
+Additionally, the sparsity of our matrix (>99.9%) suggests that many of the features will not be particularly predictive of either classification on their own.
+
+As a result, we'll attempt to reduce the dimensionality of our feature set by employing the `irlba` package to perform singular value decomposition with a target feature limit of the 300 most informative predictor combinations.
 
@@
 ```
-
-Finally, we're left with a matrix containing 40,399 features.
-While many classification models can ultimately accommodate such intense dimensionality, the computational resource requirements are quite high and would therefore make it unfeasible to run these models with the limited memory and processing power available on consumer-grade hardware.
-Additionally, the sparsity of our matrix (\>99.9%) suggests that many of the features will not be particularly predictive of either classification on their own.
-
-As a result, we'll attempt to reduce the dimensionality of our feature set by employing the `irlba` package to perform singular value decomposition with a target feature limit of the 300 most informative predictor combinations.
 
@@
 saveRDS(train.sub.irlba, file = here("results", "large-files", "train-sub-irlba.rds"))
+
+sub.svd <- as.data.frame(train.sub.irlba$v)
+sub.svd <- sub.svd %>%
+  rename_all(~ paste0("Z", seq(1, length(names(sub.svd)))))
 ```
@@
 ## Final Data Frames for Modeling
+
+Having generated structured text data that is also manageable, we can now piece together different data frames consisting of the various feature combinations we would like to assess for modeling.
+Based on the relationships observed in our exploratory data analysis those features are the token data for both email subject and body, the body length of the email, and the sender domain for the email.
 
@@
 train.sub.svd <- data.frame(label = train$label,
-                            train.sub.irlba$v)
+                            sub.svd)
 saveRDS(train.sub.svd, file = here("results", "large-files", "train-sub-svd.rds"))
@@
                                    length = train$bodylength,
-                                   train.sub.irlba$v)
+                                   sub.svd)
 saveRDS(train.sub.svd.length, file = here("results", "large-files", "train-sub-svd-length.rds"))
@@
                                           senderdomain = train$senderdomain,
-                                          train.sub.irlba$v)
+                                          sub.svd)
 saveRDS(train.sub.svd.length.sender, file = here("results", "large-files", "train-sub-svd-length-sender.rds"))
@@
                             train.body.irlba$v,
-                            train.sub.irlba$v)
+                            sub.svd)
 saveRDS(train.all.svd, file = here("results", "large-files", "train-all-svd.rds"))
@@
                                    train.body.irlba$v,
-                                   train.sub.irlba$v)
+                                   sub.svd)
 saveRDS(train.all.svd.length, file = here("results", "large-files", "train-all-svd-length.rds"))
@@
                                           train.body.irlba$v,
-                                          train.sub.irlba$v)
+                                          sub.svd)
 saveRDS(train.all.svd.length.sender, file = here("results", "large-files", "train-all-svd-length-sender.rds"))
 ```
-
-Having generated structured text data that is also manageable, we can now piece together different data frames consisting of the various feature combinations we would like to assess for modeling.
-Based on the relationships observed in our exploratory data analysis those features are the token data for both email subject and body, the body length of the email, and the sender domain for the email.
 
@@
 
+## XGBoost Iterations
+
+```{r}
+###### Body Tokens Only ######
+# tune an XGBoost model with body token SVD
+xg.body.svd <- customXG(train.body.svd)
+
+saveRDS(xg.body.svd, file = here("results", "large-files", "xg-body-svd.rds"))
+
+# tune an XGBoost model with body token SVD and email length
+xg.body.svd.length <- customXG(train.body.svd.length)
+
+saveRDS(xg.body.svd.length, file = here("results", "large-files", "xg-body-svd-length.rds"))
+
+### can't be run in a reasonable timeframe
+### too computationally intensive for a MacBook
+# tune an XGBoost model with body token SVD, email length, and sender domain
+#xg.body.svd.length.sender <- customXG(train.body.svd.length.sender)
+
+#saveRDS(xg.body.svd.length.sender, file = here("results", "large-files", "xg-body-svd-length-sender.rds"))
+
+###### Subject Tokens Only ######
+# tune an XGBoost model with subject token SVD
+xg.sub.svd <- customXG(train.sub.svd)
+
+saveRDS(xg.sub.svd, file = here("results", "large-files", "xg-sub-svd.rds"))
+
+# tune an XGBoost model with subject token SVD and email length
+xg.sub.svd.length <- customXG(train.sub.svd.length)
+
+saveRDS(xg.sub.svd.length, file = here("results", "large-files", "xg-sub-svd-length.rds"))
+
+### can't be run in a reasonable timeframe
+### too computationally intensive for a MacBook
+# tune an XGBoost model with subject token SVD, email length, and sender domain
+#xg.sub.svd.length.sender <- customXG(train.sub.svd.length.sender)
+
+#saveRDS(xg.sub.svd.length.sender, file = here("results", "large-files", "xg-sub-svd-length-sender.rds"))
+
+###### All Tokens ######
+# tune an XGBoost model with all token SVD
+xg.all.svd <- customXG(train.all.svd)
+
+saveRDS(xg.all.svd, file = here("results", "large-files", "xg-all-svd.rds"))
+
+# tune an XGBoost model with all token SVD and email length
+xg.all.svd.length <- customXG(train.all.svd.length)
+
+saveRDS(xg.all.svd.length, file = here("results", "large-files", "xg-all-svd-length.rds"))
+
+### can't be run in a reasonable timeframe
+### too computationally intensive for a MacBook
+# tune an XGBoost model with all token SVD, email length, and sender domain
+#xg.all.svd.length.sender <- customXG(train.all.svd.length.sender)
+
+#saveRDS(xg.all.svd.length.sender, file = here("results", "large-files", "xg-all-svd-length-sender.rds"))
+```
+
 # Prepping Test Data for Predictions
@@
 ## Body Tokens and SVD
+
+Having now run and saved our respective iterations of various models and features we can now proceed with preparing the test data so that it exists in the same features space with the same data transformations.
+The first step of which is to follow tokenization utilizing the same exact methods as we did for the training data.
 
@@
 
+We begin to deviate from our processing workflow utilized in the training section in order to ensure that the data is mapped into the same feature space as the training data used for modeling.
+In order to ensure this we'll utilize the `dfm_match()` function to project the same token space from the training data onto our test data.
+
 ```{r}
@@
                              tolower = FALSE)
-```
 
-```{r}
 # project the feature set from the train split onto the test split
 test.body.tokens.dfm <- dfm_match(test.body.tokens.dfm, features = featnames(train.body.tokens.dfm))
-```
 
-```{r}
 # convert dfm to matrix
@@
 
+Similarly we need to ensure the inverse document frequency values used to discount overrepresented terms are the same as those utilized in our training data when tranforming to TF-IDFs.
+
 ```{r}
 test.body.tokens.tf <- apply(test.body.tokens.matrix, 1, term.freq)
-```
 
-```{r}
 test.body.tokens.tfidf <- apply(test.body.tokens.tf, 2, tf.idf, idf = train.body.tokens.idf)
-```
 
-```{r}
 # transpose and fix missing values
@@
 ```
+
+Finally we'll be using the same SVD process as was used on the training data. We'll project the data using our custom `projSVD()` function defined in the Controls and Functions section above.
 
@@
 ## Subject Tokens and SVD
+
+*This section features the same steps as above, but now for the email subject text. Justifications for each process will be the same as before.*
 
@@
 ## Final Data Frames for Predictions
+
+Having now completed the transformations necessary to map the test data into the same space as the training data used for modeling we can proceed with making predictions and assessing model performance.
+The latter of which will be continued in a second statistical analysis `.QMD` file using the pertinent, saved `.RDS` files from this document.
 
@@
 
-# Generate Predictions
-
 ```{r}
-# predict test data with model
-#rf.body.preds <- predict(rf.body, test.body.svd)
-#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
-#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
+xg.body.svd.preds <- predict(xg.body.svd, test.body.svd)
+confusionMatrix(xg.body.svd.preds, test.body.svd$label)
+xg.all.svd.length.preds <- predict(xg.all.svd.length, test.all.svd.length)
+confusionMatrix(xg.all.svd.length.preds, test.body.svd$label)
 ```
-
-The `echo: false` option disables the printing of code (only output is displayed).

### 2024-07-18 code/analysis-code/statistical-analysis-continued.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

--- 2024-07-18 code/analysis-code/statistical-analysis.qmd
@@
 
+Having now run and saved our respective iterations of various models and features we can now proceed with preparing the test data so that it exists in the same features space with the same data transformations.
+The first step of which is to follow tokenization utilizing the same exact methods as we did for the training data.
+
 ```{r}
@@
 
-# Generate Predictions
-
-```{r}
-# predict test data with model
-#rf.body.preds <- predict(rf.body, test.body.svd)
-#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
-#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
-```
-
 The `echo: false` option disables the printing of code (only output is displayed).

### 2024-07-18 products/manuscript/Manuscript.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

@@
 title: "Fishing for Spam"
-subtitle: "Part 1: Project Proposal"
 author: Sean O'Sullivan
@@
 ```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-library(here)
-library(knitr)
+# load needed packages. make sure they are installed.
+pacman::p_load(here, #for data loading/saving
+               tidyverse, # for data manipulation and plotting
+               skimr, # for data exploration and summary
+               ggthemes) # for plot theming
 ```
@@
 The data set(s) that I'd liked to examine are a handful of annotated data sets pertaining to fraudulent (spam) emails.
-They come from a handful of sources, te largest of the bunch is from the Conference on Email and Anti-Spam.
-That data set and a pair of others found on [Kaggle](https://www.kaggle.com/datasets/naserabdullahalam/phishing-email-dataset/data?select=CEAS_08.csv) contain text information from the subjects and bodies of the emails as well as the senders, receivers, timestamps, and whether or not URLs are contained within.
+They come from a handful of sources, the largest of the bunch is from the Conference on Email and Anti-Spam.
+That data set and a pair of others can be found on [Kaggle](https://www.kaggle.com/datasets/naserabdullahalam/phishing-email-dataset/data?select=CEAS_08.csv) contain text information from the subjects and bodies of the emails as well as the senders, receivers, timestamps, and whether or not URLs are contained within.
 
@@
 I'd like to explore the features that are prevalent in the identification of spam emails.
-Utilizing text analysis in addition to the other email attributes supplied my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for automated detection as used by some popular email providers and large internal IT departments across the corporate landscape.
+Utilizing text analysis, in addition to the other email attributes supplied and engineered, my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for use of automated detection by some popular email providers and large internal IT departments across the corporate landscape.
 
@@
 
-![](../../results/figures/bodylength-distribution.png){.lightbox} First examining the distribution of email body length between spam and legitimate emails we observe a tendency for spam emails to be shorter in length.
+First examining the distribution of email body length between spam and legitimate emails we observe a tendency for spam emails to be shorter in length.
 However, interestingly the distribution of spam emails is bi-modal -- with a second peak at around the 4,000 character count -- which may indicate that spam vendors are aware of the identification potential of email length and have begun targeting a longer length to avoid notice.
+![](../../results/figures/bodylength-distribution.png){.lightbox}
 
-![](../../results/figures/subjectlength-distribution.png){.lightbox} Similarly, when we examine subject line length we observe the same tendency for spam emails to trend shorter.
+Similarly, when we examine subject line length we observe the same tendency for spam emails to trend shorter.
 Subject line does not share the same distinct bi-modal shape for spam emails that the bodies do, but subject line length distribution between both spam and legitimate emails overlap in large part which may reduce it's efficacy for classification modeling.
+![](../../results/figures/subjectlength-distribution.png){.lightbox}
 
-::: {layout-ncol="2"}
-![](../../results/figures/spam-by-domain.png)
+When plotting body tokens we can see that 'http', which precedes most links within each email, is the most common token for each label.
+Considering this token should approximate not just the presence of a URL, but also the quantity of URLs within each individual email, it will be interesting to see if including the supplied `URLs` boolean will be of value or not in our end model.
+Given the below, we suspect that the overlap will result in unnecessary noise and dimensionality to our final model.
+Additionally, considering the above bar chart suggests link presence and quantity is very similarly distributed across both labels, it may wind up improving model performance altogether to exclude 'http' or URL features from the final model.
 
-![](../../results/figures/legit-by-domain.png)
-:::
+![](../../results/figures/popular-body-tokens.png)
 
@@
 This bodes well for the classification power of sender domain when we enter the modeling phase and begin to incorporate bag-of-words and text processing.
+
+![](../../results/figures/popular-domains.png)
 

### 2024-07-18 code/analysis-code/statistical-analysis.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

@@
 
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
 # tune a Random Forest model with body token SVD, email length, and sender domain
-#rf.body.svd.length.sender <- customRF(train.body.svd.length.sender)
+rf.body.svd.length.sender <- customRF(train.body.svd.length.sender)
 
-#saveRDS(rf.body.svd.length.sender, file = here("results", "large-files", "rf-body-svd-length-sender.rds"))
+saveRDS(rf.body.svd.length.sender, file = here("results", "large-files", "rf-body-svd-length-sender.rds"))
 
@@
 
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
 # tune a Random Forest model with subject token SVD, email length, and sender domain
-#rf.sub.svd.length.sender <- customRF(train.sub.svd.length.sender)
+rf.sub.svd.length.sender <- customRF(train.sub.svd.length.sender)
 
-#saveRDS(rf.sub.svd.length.sender, file = here("results", "large-files", "rf-sub-svd-length-sender.rds"))
+saveRDS(rf.sub.svd.length.sender, file = here("results", "large-files", "rf-sub-svd-length-sender.rds"))
 
@@
 
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
 # tune a Random Forest model with all token SVD, email length, and sender domain
-#rf.all.svd.length.sender <- customRF(train.all.svd.length.sender)
+rf.all.svd.length.sender <- customRF(train.all.svd.length.sender)
 
-#saveRDS(rf.all.svd.length.sender, file = here("results", "large-files", "rf-all-svd-length-sender.rds"))
+saveRDS(rf.all.svd.length.sender, file = here("results", "large-files", "rf-all-svd-length-sender.rds"))
 ```
@@
 
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
 # tune an SVM model with body token SVD, email length, and sender domain
-#svm.body.svd.length.sender <- customSVM(train.body.svd.length.sender)
+svm.body.svd.length.sender <- customSVM(train.body.svd.length.sender)
 
-#saveRDS(svm.body.svd.length.sender, file = here("results", "large-files", "svm-body-svd-length-sender.rds"))
+saveRDS(svm.body.svd.length.sender, file = here("results", "large-files", "svm-body-svd-length-sender.rds"))
 
@@
 
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
 # tune an SVM model with subject token SVD, email length, and sender domain
-#svm.sub.svd.length.sender <- customSVM(train.sub.svd.length.sender)
+svm.sub.svd.length.sender <- customSVM(train.sub.svd.length.sender)
 
-#saveRDS(svm.sub.svd.length.sender, file = here("results", "large-files", "svm-sub-svd-length-sender.rds"))
+saveRDS(svm.sub.svd.length.sender, file = here("results", "large-files", "svm-sub-svd-length-sender.rds"))
 
@@
 
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
 # tune an SVM model with all token SVD, email length, and sender domain
-#svm.all.svd.length.sender <- customSVM(train.all.svd.length.sender)
+svm.all.svd.length.sender <- customSVM(train.all.svd.length.sender)
 
-#saveRDS(svm.all.svd.length.sender, file = here("results", "large-files", "svm-all-svd-length-sender.rds"))
+saveRDS(svm.all.svd.length.sender, file = here("results", "large-files", "svm-all-svd-length-sender.rds"))
 ```
@@
 # predict test data with model
-#rf.body.preds <- predict(rf.body, test.body.svd)
-#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
-#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
+rf.body.preds <- predict(rf.body, test.body.svd)
+rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
+rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
 ```

### 2024-07-18 code/eda-code/eda.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

@@
 ```{r}
-#load needed packages. make sure they are installed.
-library(here) #for data loading/saving
-library(dplyr)
-library(skimr)
-library(ggplot2)
-library(forcats)
-library(plyr)
-library(ggthemes)
+# load needed packages. make sure they are installed.
+pacman::p_load(here, #for data loading/saving
+               tidyverse, # for data manipulation and plotting
+               skimr, # for data exploration and summary
+               ggthemes, # for plot theming
+               tidytext) # for text analysis
 ```
@@
 ```{r}
-#Path to data. Note the use of the here() package and not absolute paths
+# path to data
 data_location <- here::here("data","processed-data","processeddata.rds")
-#load data
+# load data
 mydata <- readRDS(data_location)
-mydata$label = fct_recode(mydata$label, "No" = "0", "Yes" = "1")
 ```
@@
 
-```{r}
+```{r, warning=FALSE}
 p1 <- mydata %>% ggplot(aes(x = bodylength, fill = label)) +
@@
       xlim(0,7500) + xlab("Length") + ylab("Density") +
-      labs(title = "Body Distribution", caption = "x-axis limited to a maximum of 7,500", fill = "Spam") +
+      labs(title = "Body Distribution", caption = "x-axis limited to a maximum of 7,500", fill = NULL) +
+      scale_y_continuous(labels = scales::percent) +
       theme_tufte() +
@@
 figure_file = here::here("results", "figures","bodylength-distribution.png")
-ggsave(filename = figure_file, plot=p1) 
+ggsave(filename = figure_file, plot = p1) 
 ```
@@
 
-```{r}
+```{r, warning=FALSE}
 p2 <- mydata %>% ggplot(aes(x=subjectlength, fill = label)) +
@@
       xlim(0,150) + xlab("Length") + ylab("Density") +
-      labs(title = "Subject Line Distribution", caption = "x-axis limited to a maximum of 150", fill = "Spam") +
+      labs(title = "Subject Line Distribution", caption = "x-axis limited to a maximum of 150", fill = NULL) +
+      scale_y_continuous(labels = scales::percent) +
       theme_tufte() +
@@
 figure_file = here::here("results", "figures","subjectlength-distribution.png")
-ggsave(filename = figure_file, plot=p2) 
+ggsave(filename = figure_file, plot = p2) 
 ```
-Now the most common sender domains for spam emails. We'll continue to utilize `ggplot2`, this time with the `geom_col` shape in order to produce a bar chart of counts.
+Now the most common sender domains by both spam and legitimate emails.
+We'll continue to utilize `ggplot2`, this time with the `geom_col` shape in order to produce a bar chart of counts faceted by the label.
 
 ```{r}
-p3 <- mydata[which(mydata$label == "Yes" & !(mydata$senderdomain == "")),] %>%
-      dplyr::count(senderdomain) %>%
-      arrange(desc(n)) %>%
-      head(25) %>%
-      ggplot(aes(x = n, y = fct_reorder(senderdomain, n))) +
-      geom_col(fill = "turquoise3", colour = "black", alpha = 0.65) +
-      labs(title = "Spam Emails by Sender Domain") +
-      ylab("") + xlab("Count") +
+p3 <- mydata %>% 
+        select(label, senderdomain) %>% 
+        filter(!senderdomain == "") %>% 
+        count(label, senderdomain) %>% 
+        group_by(label) %>% 
+        slice_max(n, n = 10) %>% 
+        ungroup() %>% 
+        mutate(senderdomain = reorder(senderdomain, n)) %>% 
+        ggplot(aes(n, senderdomain, fill = label)) +
+        geom_col() +
+        facet_wrap(~label, scales = "free") +
+        labs(title = "Most Popular Sender Domains by Type") +
+        ylab(NULL) + xlab(NULL) +
+        scale_x_continuous(labels = scales::comma) +
         theme_tufte() +
@@
 plot(p3)
-figure_file = here::here("results", "figures", "spam-by-domain.png")
-ggsave(filename = figure_file, plot=p3) 
+figure_file = here::here("results", "figures", "popular-domains.png")
+ggsave(filename = figure_file, plot = p3) 
 ```
-And similarly we'll produce a bar chart for legitimate emails.
 
 ```{r}
-p3 <- mydata[which(mydata$label == "No" & !(mydata$senderdomain == "")),] %>%
-      dplyr::count(senderdomain) %>%
-      arrange(desc(n)) %>%
-      head(25) %>%
-      ggplot(aes(x = n, y = fct_reorder(senderdomain, n))) +
-      geom_col(fill = "red3", colour = "black", alpha = 0.35) +
-      labs(title = "Legit Emails by Sender Domain") +
-      ylab("") + xlab("Count") +
+# plot the most popular email body tokens
+p4 <- mydata %>%
+        select(label, subject) %>% 
+        unnest_tokens(word, subject) %>% 
+        anti_join(stop_words) %>% 
+        filter(!grepl('[0-9]', word)) %>% 
+        count(label, word) %>% 
+        group_by(label) %>% 
+        slice_max(n, n = 15) %>% 
+        ungroup() %>% 
+        mutate(word = reorder(word, n)) %>% 
+        ggplot(aes(n, word, fill = label)) +
+        geom_col() +
+        facet_wrap(~label, scales = "free") +
+        labs(title = "Most Popular Subject Tokens by Type") +
+        ylab(NULL) + xlab(NULL) +
+        scale_x_continuous(labels = scales::comma) +
         theme_tufte() +
         theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.position = "none")
-plot(p3)
-figure_file = here::here("results", "figures", "legit-by-domain.png")
-ggsave(filename = figure_file, plot=p3) 
+plot(p4)
+figure_file = here::here("results", "figures", "popular-subject-tokens.png")
+ggsave(filename = figure_file, plot = p4)
+
+p5 <- mydata %>%
+        select(label, body) %>% 
+        unnest_tokens(word, body) %>% 
+        anti_join(stop_words) %>% 
+        filter(!grepl('[0-9]', word)) %>% 
+        count(label, word) %>% 
+        group_by(label) %>% 
+        slice_max(n, n = 15) %>% 
+        ungroup() %>% 
+        mutate(word = reorder(word, n)) %>% 
+        ggplot(aes(n, word, fill = label)) +
+        geom_col() +
+        facet_wrap(~label, scales = "free") +
+        labs(title = "Most Popular Body Tokens by Type") +
+        ylab(NULL) + xlab(NULL) +
+        scale_x_continuous(labels = scales::comma) +
+        theme_tufte() +
+        theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.position = "none")
+plot(p5)
+figure_file = here::here("results", "figures", "popular-body-tokens.png")
+ggsave(filename = figure_file, plot = p5) 
 ```
 
-# Notes
+Similarly we can plot the most common body tokens and the most common subject tokens to get an idea of the word choice between the two labels.
 
-For your own explorations, tables and figures can be "quick and dirty".
-As long as you can see what's going on, there is no need to polish them.
-That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.).
-Those should look as nice, polished and easy to understand as possible.
+When plotting subject tokens we can see completely distinct collections of the most popular tokens across labels.
+This bodes well for subject tokens as a potential feature component to supplement analysis of the email bodies with in our final model.
+
+When plotting body tokens we can see that 'http', which precedes most links within each email, is the most common token for each label.
+Considering this token should approximate not just the presence of a URL, but also the quantity of URLs within each individual email, it will be interesting to see if including the supplied `URLs` boolean will be of value or not in our end model.
+We suspect that the overlap will result in unnecessary noise and dimensionality to our final model.
+Additionally, considering the above bar chart suggests link presence and quantity is very similarly distributed across both labels, it may wind up improving model performance altogether to exclude 'http' or URL features from the final model.
+
+```{r, warning=FALSE}
+p6 <- mydata %>% 
+        select(label, recursive) %>% 
+        count(label, recursive) %>% 
+        group_by(label) %>% 
+        summarize(percent = n / sum(n)) %>% 
+        ungroup() %>% 
+        filter(percent < .5) %>% 
+        ggplot(aes(label, percent, fill = label)) +
+        geom_col() +
+        labs(title = "Emails with the Same Sender/Receiver") +
+        ylab(NULL) + xlab(NULL) +
+        scale_y_continuous(labels = scales::percent) +
+        theme_tufte() +
+        theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.position = "none")
+plot(p6)
+figure_file = here::here("results", "figures", "recursive-emails.png")
+ggsave(filename = figure_file, plot = p6) 
+```
+
+Lastly, we examine an engineered feature called `recursive`.
+This feature is a boolean for whether or not the sender and receiver emails are the same.
+A common strategy when sending emails in bulk is to include the sender email in the To line while hiding a large list of recipients within the BCC line.
+In order to see if this is a strategy employed significantly by spam email senders we plotted the percentage of recursive emails for both legitimate and spam emails.
+We can see that in both cases the percentages are quite low (<2.5%) with only about a 0.7% difference between the two labels.
+Based on this, it seems unlikely that this feature will warrant further investigation or inclusion in our final model.

### 2024-07-18 code/processing-code/processingcode.R

@@ -12,12 +12,10 @@
 
 ## ---- packages --------
 # load needed packages. make sure they are installed.
-library(dplyr) # for data processing/cleaning
-library(tidyverse) # for data processing/cleaning
-library(skimr) # for nice visualization of data 
-library(here) # to set paths
-library(lubridate) # to convert datetimes
-library(stringr) # to perform regex extractions
+pacman::p_load(tidyverse, # for data processing/cleaning
+               skimr, # for nice visualization of data
+               here, # to set paths
+               lubridate) # to convert datetimes
 
 
 ## ---- loaddata --------
@@ -70,19 +68,22 @@ processeddata <- processeddata %>%
 # some sender and receiver emails either lack an email address or lack any attached contact name to be removed
 # we utilize the dplyr 'if_else' function to handle separate cases and ensure we're returning consistent
 # results throughout all rows of the dataset
-processeddata <- processeddata %>% mutate(
-                                          recursive = processeddata$sender == processeddata$receiver,
-                                          senderdomain = if_else(
-                                                                is.na(str_extract(processeddata$sender, "(?<=@).*")),
-                                                                "",
-                                                                str_extract(processeddata$sender, "(?<=@).*")
-                                                                ),
-                                          receiverdomain = if_else(
-                                                                  is.na(str_extract(processeddata$receiver, "(?<=@).*")),
-                                                                  "",
-                                                                  str_extract(processeddata$receiver, "(?<=@).*")
-                                                                  ),
-                                          )
+processeddata <- processeddata %>% 
+                  mutate(
+                        recursive = processeddata$sender == processeddata$receiver,
+                        senderdomain = if_else(
+                                              is.na(str_extract(processeddata$sender, "(?<=@).*")),
+                                              "",
+                                              str_extract(processeddata$sender, "(?<=@).*")
+                                              ),
+                        receiverdomain = if_else(
+                                                is.na(str_extract(processeddata$receiver, "(?<=@).*")),
+                                                "",
+                                                str_extract(processeddata$receiver, "(?<=@).*")
+                                                )) %>% 
+                  mutate(senderdomain = as.factor(senderdomain),
+                         receiverdomain = as.factor(receiverdomain))
+processeddata$label <- fct_recode(processeddata$label, "Legitimate" = "0", "Spam" = "1")
 skimr::skim(processeddata)
 
 ## ---- savedata --------

### 2024-07-18 products/manuscript/Manuscript.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

@@
 title: "Fishing for Spam"
+subtitle: "Part 1: Project Proposal"
 author: Sean O'Sullivan
@@
 I'd like to explore the features that are prevalent in the identification of spam emails.
-Utilizing text analysis, in addition to the other email attributes supplied and engineered, my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for use of automated detection by some popular email providers and large internal IT departments across the corporate landscape.
+Utilizing text analysis in addition to the other email attributes supplied my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for automated detection as used by some popular email providers and large internal IT departments across the corporate landscape.
 

### 2024-08-08 code/analysis-code/statistical-analysis-continued.qmd
Final commit

--- 2024-07-18 code/analysis-code/statistical-analysis.qmd
@@
 ---
-title: "Statistical Analysis"
+title: "Statistical Analysis Continued"
 format: html
@@
 
-# Prepping Train Data for Modeling
+# Model Evaluation
 
@@
 
-Our first step is to import our processed data set and load in all necessary packages and functions for the analysis.
+Our first step is to import our model data and load in all necessary test data frames for evaluating model success.
 Additionally we'll set the seed for the entirety of this document for the purposes of ensuring the exact values of this analysis are reproducible.
@@
                randomForest, # for model training
+               gridExtra,
                irlba, # for SVD
+               pROC, # for ROC curves
                doSNOW, # for parallel processing
@@
 
-# path to data
-data_location <- here::here("data","processed-data","processeddata.rds")
-# load data
-mydata <- readRDS(data_location)
+# load model data and test data frames
+# Random Forest Models
+rf.body.svd <- readRDS(here("results", "large-files", "rf-body-svd.rds"))
+rf.body.svd.length <- readRDS(here("results", "large-files", "rf-body-svd-length.rds"))
+rf.sub.svd <- readRDS(here("results", "large-files", "rf-sub-svd.rds"))
+rf.sub.svd.length <- readRDS(here("results", "large-files", "rf-sub-svd-length.rds"))
+rf.all.svd <- readRDS(here("results", "large-files", "rf-all-svd.rds"))
+rf.all.svd.length <- readRDS(here("results", "large-files", "rf-all-svd-length.rds"))
+
+# SVM Models
+svm.body.svd <- readRDS(here("results", "large-files", "svm-body-svd.rds"))
+svm.body.svd.length <- readRDS(here("results", "large-files", "svm-body-svd-length.rds"))
+svm.sub.svd <- readRDS(here("results", "large-files", "svm-sub-svd.rds"))
+svm.sub.svd.length <- readRDS(here("results", "large-files", "svm-sub-svd-length.rds"))
+svm.all.svd <- readRDS(here("results", "large-files", "svm-all-svd.rds"))
+svm.all.svd.length <- readRDS(here("results", "large-files", "svm-all-svd-length.rds"))
+
+# XGBoost Models
+xg.body.svd <- readRDS(here("results", "large-files", "xg-body-svd.rds"))
+xg.body.svd.length <- readRDS(here("results", "large-files", "xg-body-svd-length.rds"))
+xg.sub.svd <- readRDS(here("results", "large-files", "xg-sub-svd.rds"))
+xg.sub.svd.length <- readRDS(here("results", "large-files", "xg-sub-svd-length.rds"))
+xg.all.svd <- readRDS(here("results", "large-files", "xg-all-svd.rds"))
+xg.all.svd.length <- readRDS(here("results", "large-files", "xg-all-svd-length.rds"))
+
+# test data frames
+test.body.svd <- readRDS(here("results", "large-files", "test-body-svd.rds"))
+test.body.svd.length <- readRDS(here("results", "large-files", "test-body-svd-length.rds"))
+test.sub.svd <- readRDS(here("results", "large-files", "test-sub-svd.rds"))
+test.sub.svd.length <- readRDS(here("results", "large-files", "test-sub-svd-length.rds"))
+test.all.svd <- readRDS(here("results", "large-files", "test-all-svd.rds"))
+test.all.svd.length <- readRDS(here("results", "large-files", "test-all-svd-length.rds"))
 
@@
 
-## Train/Test Splits
+## Generating Predictions on the Test Set
 
-Next we'll focus on splitting the data into train and test splits.
-We'll opt to create a partition that leaves us 30% of the processed data set for testing the generalizability of our final model.
-We'll also want to ensure that class representations are maintained proportionate to the original data set.
+Then we'll create predictions for the test data with our tuned models and generate confusion matrices for each to evaluate the overall performance of each.
+We'll use this information to help inform which models generalized best when classifying new data and will assess which types of mistakes the models tend to err on the side of -- for our purposes allowing the occasional spam message through is likely better than overzealously blocking legitimate emails from reaching the user's inbox.
+
+In order to determine the models least likely of overzealously blocking legitimate emails we'll consider evaluating model performance with either Recall or F1 score The former ensures we're classifying as many true legitimate emails correctly as possible, while the latter assesses the overall accuracy of our classification while factoring in the weights of any class imbalance.
 
 ```{r}
-# create smaller partition for the sake of being able to run this shit on my PC
-subsample <- createDataPartition(mydata$label,
-                                 times = 1,
-                                 p = .1,
-                                 list = FALSE)
+# create all test predictions
+# Random Forest Models
+rf.body.svd.preds <- predict(rf.body.svd, test.body.svd)
+rf.body.svd.length.preds <- predict(rf.body.svd.length, test.body.svd.length)
+rf.sub.svd.preds <- predict(rf.sub.svd, test.sub.svd)
+rf.sub.svd.length.preds <- predict(rf.sub.svd.length, test.sub.svd.length)
+rf.all.svd.preds <- predict(rf.all.svd, test.all.svd)
+rf.all.svd.length.preds <- predict(rf.all.svd.length, test.all.svd.length)
 
-sub.data <- mydata[subsample,]
+# SVM Models
+svm.body.svd.preds <- predict(svm.body.svd, test.body.svd)
+svm.body.svd.length.preds <- predict(svm.body.svd.length, test.body.svd.length)
+svm.sub.svd.preds <- predict(svm.sub.svd, test.sub.svd)
+svm.sub.svd.length.preds <- predict(svm.sub.svd.length, test.sub.svd.length)
+svm.all.svd.preds <- predict(svm.all.svd, test.all.svd)
+svm.all.svd.length.preds <- predict(svm.all.svd.length, test.all.svd.length)
 
-# create index of partition
-inTrain <- createDataPartition(sub.data$label, 
-                               times = 1, 
-                               p = 0.7,
-                               list = FALSE)
-
-# subset training data
-train <- sub.data[inTrain,]
-
-# subset test data
-test <- sub.data[-inTrain,]
-```
-
-The output of our class proportions look correctly maintained by the `createDataPartition()` function so we can proceed with preparing our training data for future models.
-
-## Controls and Functions
-
-```{r}
-# create CV folds and model controls
-folds <- createMultiFolds(train$label,
-                          k = 10,
-                          times = 3)
-ctrl <- trainControl(method = "repeatedcv",
-                     number = 10,
-                     repeats = 3,
-                     index = folds)
-gridSVM <- expand.grid(C = seq(.1, 1, by = .1),
-                       sigma = c(0.01, 0.1, 0.5, 1))
-
-# function definitions
-# calculate term frequency
-term.freq <- function(row) {
-  row / sum(row)
-}
-
-# calculate inverse document frequency
-inverse.doc.freq <- function(col) {
-  corpus.size <- length(col)
-  doc.count <- length(which(col > 0))
-
-  log10(corpus.size / doc.count)
-}
-
-# calculate tf-idf
-tf.idf <- function(tf, idf) {
-  tf * idf
-}
-
-# project singular value decomp onto test split
-projSVD <- function(trainirlba, testtfidf) {
-  sigma.inverse <- 1 / trainirlba$d
-  u.transpose <- t(trainirlba$u)
-  projection <- t(sigma.inverse * u.transpose %*% t(testtfidf))
-  projection <- as.data.frame(projection)
-  names(projection) <- paste0("X", seq(1,300))
-  return(projection)
-}
-
-# tune Random Forest model
-customRF <- function(df) {
-  cores <- detectCores()
-  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
-                                                   # of cores for parallel processing
-  registerDoParallel(clus)
-  model <- train(label ~ ., 
-                 data = df,
-                 method = "rf",
-                 trControl = ctrl,
-                 tuneLength = 7)
-  stopCluster(clus)
-  return(model)
-}
-
-# tune an SVM model
-customSVM <- function(df) {
-  cores <- detectCores()
-  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
-                                                   # of cores for parallel processing
-  registerDoParallel(clus)
-  model <- train(label ~ ., 
-                 data = df,
-                 method = "svmRadial",
-                 trControl = ctrl,
-                 tuneLength = 7,
-                 tuneGrid = gridSVM)
-  stopCluster(clus)
-  return(model)
-}
-```
-
-To reduce repetitive model configurations and parameter tuning grids we'll establish some variables and functions for later use in the analysis.
-
-## Body Tokens and SVD
-
-```{r}
-# generate tokens from email bodies
-train.body.tokens <- tokens(train$body, what = "word",
-                       remove_numbers = TRUE,
-                       remove_separators = TRUE,
-                       remove_punct = TRUE,
-                       remove_symbols = TRUE,
-                       split_hyphens = FALSE,
-                       split_tags = TRUE,
-                       remove_url = FALSE)
-```
-
-Our first step in structuring text data requires that we tokenize the email bodies.
-
-```{r}
-# lowercase tokens
-train.body.tokens <- tokens_tolower(train.body.tokens)
-
-# remove stopwords
-train.body.tokens <- tokens_select(train.body.tokens,
-                                   stopwords(),
-                                   selection = "remove")
-
-# stem words
-train.body.tokens <- tokens_wordstem(train.body.tokens,
-                                     language = "english")
-```
-
-After which we'll proceed with standardizing and consolidating the tokens.
-In our case we'll be lowercasing all tokens, removing English stopwords, and consolidating the semantics by stemming the remaining tokens.
-
-```{r}
-# create document-feature matrix
-train.body.tokens.dfm <- dfm(train.body.tokens,
-                             tolower = FALSE)
-
-# convert dfm to matrix
-train.body.tokens.matrix <- as.matrix(train.body.tokens.dfm)
-```
-
-After we've tokenized we're now able to restructure the data as a document-feature matrix in preparation for calculating term frequencies and inverse document frequencies.
-
-```{r}
-# create the term frequency
-train.body.tokens.tf <- apply(train.body.tokens.matrix, 1, term.freq)
-
-# create the inverse document frequency
-train.body.tokens.idf <- apply(train.body.tokens.matrix, 2, inverse.doc.freq)
-
-# create the tf-idf
-train.body.tokens.tfidf <- apply(train.body.tokens.tf, 2, tf.idf, idf = train.body.tokens.idf)
-```
-
-We proceed by calculating TF-IDFs for the email body token data.
-This is valuable because it standardizes the word occurrences and further helps to reduce the influence of over represented words on the model.
-The TF values will accomplish the former while the IDF values will accomplish the latter by discounting the tokens found consistently throughout the corpus of emails.
-In the case of tokens found in every email throughout the corpus the end result is effectively complete removal from the model akin to our earlier removal of stop words.
-
-```{r}
-# transpose tf-idf and handle incomplete cases
-train.body.tokens.tfidf <- t(train.body.tokens.tfidf)
-incomplete.cases <- which(!complete.cases(train.body.tokens.tfidf))
-train.body.tokens.tfidf[incomplete.cases,] <- rep(0.0, ncol(train.body.tokens.tfidf))
-```
-
-Having reinterpreted our tokens as TF-IDFs we can transpose the matrix back to it's original orientation and fill in zeroes for any tokens that generated NA values (as a result of no representation of the given tokens within the email).
-
-```{r}
-# run singular value decomposition for body tokens
-train.body.irlba <- irlba(t(train.body.tokens.tfidf), nv = 300, maxit = 600)
-
-saveRDS(train.body.irlba, file = here("results", "large-files", "train-body-irlba.rds"))
-```
-
-Finally, we're left with a matrix containing 40,399 features.
-While many classification models can ultimately accommodate such intense dimensionality, the computational resource requirements are quite high and would therefore make it unfeasible to run these models with the limited memory and processing power available on consumer-grade hardware.
-Additionally, the sparsity of our matrix (\>99.9%) suggests that many of the features will not be particularly predictive of either classification on their own.
-
-As a result, we'll attempt to reduce the dimensionality of our feature set by employing the `irlba` package to perform singular value decomposition with a target feature limit of the 300 most informative predictor combinations.
-
-## Subject Tokens and SVD
-
-*This section features the same steps as above, but now for the email subject text. Justifications for each process will be the same as before.*
-
-```{r}
-# generate tokens from email subjects
-train.sub.tokens <- tokens(train$subject, what = "word",
-                       remove_numbers = TRUE,
-                       remove_separators = TRUE,
-                       remove_punct = TRUE,
-                       remove_symbols = TRUE,
-                       split_hyphens = FALSE,
-                       split_tags = TRUE,
-                       remove_url = FALSE)
+# XGBoost Models
+xg.body.svd.preds <- predict(xg.body.svd, test.body.svd)
+xg.body.svd.length.preds <- predict(xg.body.svd.length, test.body.svd.length)
+xg.sub.svd.preds <- predict(xg.sub.svd, test.sub.svd)
+xg.sub.svd.length.preds <- predict(xg.sub.svd.length, test.sub.svd.length)
+xg.all.svd.preds <- predict(xg.all.svd, test.all.svd)
+xg.all.svd.length.preds <- predict(xg.all.svd.length, test.all.svd.length)
 ```
@@
 ```{r}
-# lowercase tokens
-train.sub.tokens <- tokens_tolower(train.sub.tokens)
+# generate confusion matrices
+# Random Forest Models
+rf.body.svd.cm <- confusionMatrix(rf.body.svd.preds, test.body.svd$label)
+rf.body.svd.length.cm <- confusionMatrix(rf.body.svd.length.preds, test.body.svd.length$label)
+rf.sub.svd.cm <- confusionMatrix(rf.sub.svd.preds, test.sub.svd$label)
+rf.sub.svd.length.cm <- confusionMatrix(rf.sub.svd.length.preds, test.sub.svd.length$label)
+rf.all.svd.cm <- confusionMatrix(rf.all.svd.preds, test.all.svd$label)
+rf.all.svd.length.cm <- confusionMatrix(rf.all.svd.length.preds, test.all.svd.length$label)
 
-# remove stopwords
-train.sub.tokens <- tokens_select(train.sub.tokens,
-                                   stopwords(),
-                                   selection = "remove")
+# SVM Models
+svm.body.svd.cm <- confusionMatrix(svm.body.svd.preds, test.body.svd$label)
+svm.body.svd.length.cm <- confusionMatrix(svm.body.svd.length.preds, test.body.svd.length$label)
+svm.sub.svd.cm <- confusionMatrix(svm.sub.svd.preds, test.sub.svd$label)
+svm.sub.svd.length.cm <- confusionMatrix(svm.sub.svd.length.preds, test.sub.svd.length$label)
+svm.all.svd.cm <- confusionMatrix(svm.all.svd.preds, test.all.svd$label)
+svm.all.svd.length.cm <- confusionMatrix(svm.all.svd.length.preds, test.all.svd.length$label)
 
-# stem words
-train.sub.tokens <- tokens_wordstem(train.sub.tokens,
-                                     language = "english")
+# XGBoost Models
+xg.body.svd.cm <- confusionMatrix(xg.body.svd.preds, test.body.svd$label)
+xg.body.svd.length.cm <- confusionMatrix(xg.body.svd.length.preds, test.body.svd.length$label)
+xg.sub.svd.cm <- confusionMatrix(xg.sub.svd.preds, test.sub.svd$label)
+xg.sub.svd.length.cm <- confusionMatrix(xg.sub.svd.length.preds, test.sub.svd.length$label)
+xg.all.svd.cm <- confusionMatrix(xg.all.svd.preds, test.all.svd$label)
+xg.all.svd.length.cm <- confusionMatrix(xg.all.svd.length.preds, test.all.svd.length$label)
+
+# create a data frame of combined confusion metrics for each model
+cm.df <- rbind(rf.body.svd = rf.body.svd.cm$byClass,
+               rf.body.svd.length = rf.body.svd.length.cm$byClass,
+               rf.sub.svd = rf.sub.svd.cm$byClass,
+               rf.sub.svd.length = rf.sub.svd.length.cm$byClass,
+               rf.all.svd = rf.all.svd.cm$byClass,
+               rf.all.svd.length = rf.all.svd.length.cm$byClass,
+               svm.body.svd = svm.body.svd.cm$byClass,
+               svm.body.svd.length = svm.body.svd.length.cm$byClass,
+               svm.sub.svd = svm.sub.svd.cm$byClass,
+               svm.sub.svd.length = svm.sub.svd.length.cm$byClass,
+               svm.all.svd = svm.all.svd.cm$byClass,
+               svm.all.svd.length = svm.all.svd.length.cm$byClass,
+               xg.body.svd = xg.body.svd.cm$byClass,
+               xg.body.svd.length = xg.body.svd.length.cm$byClass,
+               xg.sub.svd = xg.sub.svd.cm$byClass,
+               xg.sub.svd.length = xg.sub.svd.length.cm$byClass,
+               xg.all.svd = xg.all.svd.cm$byClass,
+               xg.all.svd.length = xg.all.svd.length.cm$byClass)
+cm.df <- as.data.frame(cm.df) %>% 
+          rownames_to_column(var = "Model")
+saveRDS(cm.df, here("results", "tables", "confusion-matrix-compared.rds"))
 ```
 
+Taking a look at the model permutation with the best F1 score and the best Recall we can see that in both cases they are those models utilizing only email body tokens for predictive power.
+Between the two models the XGBoost model manages a more impressive F1 score and a nearly identical Recall as the radial SVM model.
+As a result we'll consider the XGBoost model with only email body tokens our winner.
+
 ```{r}
-# create document-feature matrix
-train.sub.tokens.dfm <- dfm(train.sub.tokens,
-                             tolower = FALSE)
+# plot F1 scores
+f1Plot <- cm.df %>% 
+            mutate(Model = as.factor(Model),
+                   F1 = as.numeric(F1)) %>% 
+            ggplot(aes(F1, fct_reorder(Model, F1))) +
+            geom_col(fill = "aquamarine3") +
+            theme_tufte() +
+            labs(title = "F1 Scores Compared") +
+            ylab(NULL) + xlab("F1 Score") +
+            theme(plot.title = element_text(hjust = 0.5, size = 18))
+f1Plot
+
+# plot Recall scores
+recallPlot <- cm.df %>% 
+            mutate(Model = as.factor(Model),
+                   F1 = as.numeric(Recall)) %>% 
+            ggplot(aes(F1, fct_reorder(Model, Recall))) +
+            geom_col(fill = "aquamarine3") +
+            theme_tufte() +
+            labs(title = "Recall Compared") +
+            ylab(NULL) + xlab("Recall") +
+            theme(plot.title = element_text(hjust = 0.5, size = 18))
+recallPlot
+
+# model with the highest F1
+cm.df %>% filter(F1 == max(F1))
+# model with the highest recall
+cm.df %>% filter(Recall == max(Recall))
 ```
 
-```{r}
-# convert dfm to matrix
-train.sub.tokens.matrix <- as.matrix(train.sub.tokens.dfm)
-```
+We can now start to visualize some of the features of our best model to understand what helped it achieve such high predictive scores.
+Taking a look first at the confusion matrix for our two class model we can see that there were only 18 false negatives and only 26 false positives that would have slipped through to the user.
+
+Taking a look at variable importance we begin to see the pitfalls of SVD on model interpretability.
+While we there are sharp declines in importance after the first two singular values and the second two singular values we're unable to ascertain what actual word choices in the body of the email these singular values are comprised of.
+However, considering the swift decline in importance it may be worth attempting to remodel using this same data without SVD on more powerful hardware so that better intrepretability might be possible.
+
+We do note that none of the highest performing models utilized either email subject tokens or email body length to achieve their scores.
+It's possible that the former only worked to introduce noise to the classification models and the latter was partially encompassed in the analysis of the tokens that comprise the body's length.
 
 ```{r}
-# create the term frequency
-train.sub.tokens.tf <- apply(train.sub.tokens.matrix, 1, term.freq)
+# best tuning parameters for the best model
+best.params <- t(xg.body.svd$bestTune)
+colnames(best.params) <- "Value"
+best.params <- as.data.frame(best.params) %>%
+                rownames_to_column(var = "Parameter")
+best.params
+saveRDS(best.params, here("results", "tables", "best-params.rds"))
+
+# print confusion matrix for best model
+cm.best <- xg.body.svd.cm$table
+cm.best
+
+saveRDS(cm.best, here("results", "tables", "best-confusion-matrix.rds"))
+
+# predict probabilities for best model
+proba <- predict(xg.body.svd, test.body.svd, type = "prob")
+curveROC <- roc(test.body.svd$label, proba[, 2], levels = rev(levels(test.body.svd$label)))
+curveROC
+auc_val <- auc(curveROC)
+
+par(pty = "s")
+plot.roc(curveROC, legacy.axes = T, percent = T, xlab = "False Positive Rate", ylab = "True Positive Rate", col = "aquamarine3", lwd = 5, print.auc = T)
+
+saveRDS(curveROC, here("results", "tables", "roc.rds"))
+
+# variable importance for best model
+varImpPlot <- as.data.frame(varImp(xg.body.svd, scale = F)$importance) %>% 
+                rownames_to_column(var = "Variable") %>% 
+                mutate(Variable = as.factor(Variable)) %>% 
+                slice_max(n = 20, order_by = Overall) %>% 
+                ggplot(aes(y = Overall, x = fct_reorder(Variable, desc(Overall)))) +
+                geom_col(fill = "aquamarine3") +
+                theme_tufte() +
+                labs(title = "Variable Importance") +
+                ylab(NULL) + xlab(NULL) +
+                theme(plot.title = element_text(hjust = 0.5, size = 18))
+varImpPlot
+ggsave(here("results", "figures", "final-var-imp.png"), varImpPlot)
 ```
-
-```{r}
-# create the inverse document frequency
-train.sub.tokens.idf <- apply(train.sub.tokens.matrix, 2, inverse.doc.freq)
-```
-
-```{r}
-# create the tf-idf
-train.sub.tokens.tfidf <- apply(train.sub.tokens.tf, 2, tf.idf, idf = train.sub.tokens.idf)
-```
-
-```{r}
-# transpose tf-idf and handle incomplete cases
-train.sub.tokens.tfidf <- t(train.sub.tokens.tfidf)
-incomplete.cases <- which(!complete.cases(train.sub.tokens.tfidf))
-train.sub.tokens.tfidf[incomplete.cases,] <- rep(0.0, ncol(train.sub.tokens.tfidf))
-```
-
-```{r}
-# create single value decomp for subject tokens
-train.sub.irlba <- irlba(t(train.sub.tokens.tfidf), nv = 300, maxit = 600)
-
-saveRDS(train.sub.irlba, file = here("results", "large-files", "train-sub-irlba.rds"))
-```
-
-## Final Data Frames for Modeling
-
-```{r}
-####### Body Token Only #######
-# create revised training data frame with body token SVD
-train.body.svd <- data.frame(label = train$label,
-                             train.body.irlba$v)
-saveRDS(train.body.svd, file = here("results", "large-files", "train-body-svd.rds"))
-
-# create revised training data frame with body token SVD and email length
-train.body.svd.length <- data.frame(label = train$label,
-                               length = train$bodylength,
-                               train.body.irlba$v)
-saveRDS(train.body.svd.length, file = here("results", "large-files", "train-body-svd-length.rds"))
-
-# create revised training data frame 
-# with body token SVD, email length, and sender domain
-train.body.svd.length.sender <- data.frame(label = train$label,
-                                           length = train$bodylength,
-                                           senderdomain = train$senderdomain,
-                                           train.body.irlba$v)
-saveRDS(train.body.svd.length.sender, file = here("results", "large-files", "train-body-svd-length-sender.rds"))
-
-####### Subject Token Only #######
-# create revised training data frame with subject token SVD
-train.sub.svd <- data.frame(label = train$label,
-                            train.sub.irlba$v)
-saveRDS(train.sub.svd, file = here("results", "large-files", "train-sub-svd.rds"))
-
-# create revised training data frame with subject token SVD and email length
-train.sub.svd.length <- data.frame(label = train$label,
-                                   length = train$bodylength,
-                                   train.sub.irlba$v)
-saveRDS(train.sub.svd.length, file = here("results", "large-files", "train-sub-svd-length.rds"))
-
-# create revised training data frame 
-# with subject token SVD, email length, and sender domain
-train.sub.svd.length.sender <- data.frame(label = train$label,
-                                          length = train$bodylength,
-                                          senderdomain = train$senderdomain,
-                                          train.sub.irlba$v)
-saveRDS(train.sub.svd.length.sender, file = here("results", "large-files", "train-sub-svd-length-sender.rds"))
-
-####### All Tokens #######
-# create revised training data frame with all tokens SVD
-train.all.svd <- data.frame(label = train$label,
-                            train.body.irlba$v,
-                            train.sub.irlba$v)
-saveRDS(train.all.svd, file = here("results", "large-files", "train-all-svd.rds"))
-
-# create revised training data frame with all tokens SVD and email length
-train.all.svd.length <- data.frame(label = train$label,
-                                   length = train$bodylength,
-                                   train.body.irlba$v,
-                                   train.sub.irlba$v)
-saveRDS(train.all.svd.length, file = here("results", "large-files", "train-all-svd-length.rds"))
-
-# create revised training data frame 
-# with all tokens SVD, email length, and sender domain
-train.all.svd.length.sender <- data.frame(label = train$label,
-                                          length = train$bodylength,
-                                          senderdomain = train$senderdomain,
-                                          train.body.irlba$v,
-                                          train.sub.irlba$v)
-saveRDS(train.all.svd.length.sender, file = here("results", "large-files", "train-all-svd-length-sender.rds"))
-```
-
-Having generated structured text data that is also manageable, we can now piece together different data frames consisting of the various feature combinations we would like to assess for modeling.
-Based on the relationships observed in our exploratory data analysis those features are the token data for both email subject and body, the body length of the email, and the sender domain for the email.
-
-# Tune Models
-
-## Random Forest Iterations
-
-```{r}
-###### Body Tokens Only ######
-# tune a Random Forest model with body token SVD
-rf.body.svd <- customRF(train.body.svd)
-
-saveRDS(rf.body.svd, file = here("results", "large-files", "rf-body-svd.rds"))
-
-# tune a Random Forest model with body token SVD and email length
-rf.body.svd.length <- customRF(train.body.svd.length)
-
-saveRDS(rf.body.svd.length, file = here("results", "large-files", "rf-body-svd-length.rds"))
-
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
-# tune a Random Forest model with body token SVD, email length, and sender domain
-#rf.body.svd.length.sender <- customRF(train.body.svd.length.sender)
-
-#saveRDS(rf.body.svd.length.sender, file = here("results", "large-files", "rf-body-svd-length-sender.rds"))
-
-###### Subject Tokens Only ######
-# tune a Random Forest model with subject token SVD
-rf.sub.svd <- customRF(train.sub.svd)
-
-saveRDS(rf.sub.svd, file = here("results", "large-files", "rf-sub-svd.rds"))
-
-# tune a Random Forest model with subject token SVD and email length
-rf.sub.svd.length <- customRF(train.sub.svd.length)
-
-saveRDS(rf.sub.svd.length, file = here("results", "large-files", "rf-sub-svd-length.rds"))
-
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
-# tune a Random Forest model with subject token SVD, email length, and sender domain
-#rf.sub.svd.length.sender <- customRF(train.sub.svd.length.sender)
-
-#saveRDS(rf.sub.svd.length.sender, file = here("results", "large-files", "rf-sub-svd-length-sender.rds"))
-
-###### All Tokens ######
-# tune a Random Forest model with all token SVD
-rf.all.svd <- customRF(train.all.svd)
-
-saveRDS(rf.all.svd, file = here("results", "large-files", "rf-all-svd.rds"))
-
-# tune a Random Forest model with all token SVD and email length
-rf.all.svd.length <- customRF(train.all.svd.length)
-
-saveRDS(rf.all.svd.length, file = here("results", "large-files", "rf-all-svd-length.rds"))
-
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
-# tune a Random Forest model with all token SVD, email length, and sender domain
-#rf.all.svd.length.sender <- customRF(train.all.svd.length.sender)
-
-#saveRDS(rf.all.svd.length.sender, file = here("results", "large-files", "rf-all-svd-length-sender.rds"))
-```
-
-## SVM Iterations
-
-```{r}
-###### Body Tokens Only ######
-# tune an SVM model with body token SVD
-svm.body.svd <- customSVM(train.body.svd)
-
-saveRDS(svm.body.svd, file = here("results", "large-files", "svm-body-svd.rds"))
-
-# tune an SVM model with body token SVD and email length
-svm.body.svd.length <- customSVM(train.body.svd.length)
-
-saveRDS(svm.body.svd.length, file = here("results", "large-files", "svm-body-svd-length.rds"))
-
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
-# tune an SVM model with body token SVD, email length, and sender domain
-#svm.body.svd.length.sender <- customSVM(train.body.svd.length.sender)
-
-#saveRDS(svm.body.svd.length.sender, file = here("results", "large-files", "svm-body-svd-length-sender.rds"))
-
-###### Subject Tokens Only ######
-# tune an SVM model with subject token SVD
-svm.sub.svd <- customSVM(train.sub.svd)
-
-saveRDS(svm.sub.svd, file = here("results", "large-files", "svm-sub-svd.rds"))
-
-# tune an SVM model with subject token SVD and email length
-svm.sub.svd.length <- customSVM(train.sub.svd.length)
-
-saveRDS(svm.sub.svd.length, file = here("results", "large-files", "svm-sub-svd-length.rds"))
-
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
-# tune an SVM model with subject token SVD, email length, and sender domain
-#svm.sub.svd.length.sender <- customSVM(train.sub.svd.length.sender)
-
-#saveRDS(svm.sub.svd.length.sender, file = here("results", "large-files", "svm-sub-svd-length-sender.rds"))
-
-###### All Tokens ######
-# tune an SVM model with all token SVD
-svm.all.svd <- customSVM(train.all.svd)
-
-saveRDS(svm.all.svd, file = here("results", "large-files", "svm-all-svd.rds"))
-
-# tune an SVM model with all token SVD and email length
-svm.all.svd.length <- customSVM(train.all.svd.length)
-
-saveRDS(svm.all.svd.length, file = here("results", "large-files", "svm-all-svd-length.rds"))
-
-### can't be run in a reasonable timeframe
-### too computationally intensive for a MacBook
-# tune an SVM model with all token SVD, email length, and sender domain
-#svm.all.svd.length.sender <- customSVM(train.all.svd.length.sender)
-
-#saveRDS(svm.all.svd.length.sender, file = here("results", "large-files", "svm-all-svd-length-sender.rds"))
-```
-
-# Prepping Test Data for Predictions
-
-## Body Tokens and SVD
-
-```{r}
-# generate tokens from test email bodies
-test.body.tokens <- tokens(test$body, what = "word",
-                       remove_numbers = TRUE,
-                       remove_separators = TRUE,
-                       remove_punct = TRUE,
-                       remove_symbols = TRUE,
-                       split_hyphens = FALSE,
-                       split_tags = TRUE,
-                       remove_url = FALSE)
-```
-
-```{r}
-# lowercase tokens
-test.body.tokens <- tokens_tolower(test.body.tokens)
-
-# remove stopwords
-test.body.tokens <- tokens_select(test.body.tokens,
-                                   stopwords(),
-                                   selection = "remove")
-
-# stem words
-test.body.tokens <- tokens_wordstem(test.body.tokens,
-                                     language = "english")
-```
-
-```{r}
-# create document-feature matrix
-test.body.tokens.dfm <- dfm(test.body.tokens,
-                             tolower = FALSE)
-```
-
-```{r}
-# project the feature set from the train split onto the test split
-test.body.tokens.dfm <- dfm_match(test.body.tokens.dfm, features = featnames(train.body.tokens.dfm))
-```
-
-```{r}
-# convert dfm to matrix
-test.body.tokens.matrix <- as.matrix(test.body.tokens.dfm)
-```
-
-```{r}
-test.body.tokens.tf <- apply(test.body.tokens.matrix, 1, term.freq)
-```
-
-```{r}
-test.body.tokens.tfidf <- apply(test.body.tokens.tf, 2, tf.idf, idf = train.body.tokens.idf)
-```
-
-```{r}
-# transpose and fix missing values
-test.body.tokens.tfidf <- t(test.body.tokens.tfidf)
-test.body.tokens.tfidf[is.na(test.body.tokens.tfidf)] <- 0.0
-```
-
-```{r}
-# project the SVD from the train split onto the test split
-test.body.svd.proj <- projSVD(train.body.irlba, test.body.tokens.tfidf)
-```
-
-## Subject Tokens and SVD
-
-```{r}
-# generate tokens from test email bodies
-test.sub.tokens <- tokens(test$subject, what = "word",
-                          remove_numbers = TRUE,
-                          remove_separators = TRUE,
-                          remove_punct = TRUE,
-                          remove_symbols = TRUE,
-                          split_hyphens = FALSE,
-                          split_tags = TRUE,
-                          remove_url = FALSE)
-```
-
-```{r}
-# lowercase tokens
-test.sub.tokens <- tokens_tolower(test.sub.tokens)
-
-# remove stopwords
-test.sub.tokens <- tokens_select(test.sub.tokens,
-                                 stopwords(),
-                                 selection = "remove")
-
-# stem words
-test.sub.tokens <- tokens_wordstem(test.sub.tokens,
-                                   language = "english")
-```
-
-```{r}
-# create document-feature matrix
-test.sub.tokens.dfm <- dfm(test.sub.tokens,
-                           tolower = FALSE)
-```
-
-```{r}
-# project the feature set from the train split onto the test split
-test.sub.tokens.dfm <- dfm_match(test.sub.tokens.dfm, features = featnames(train.sub.tokens.dfm))
-```
-
-```{r}
-# convert dfm to matrix
-test.sub.tokens.matrix <- as.matrix(test.sub.tokens.dfm)
-```
-
-```{r}
-test.sub.tokens.tf <- apply(test.sub.tokens.matrix, 1, term.freq)
-```
-
-```{r}
-test.sub.tokens.tfidf <- apply(test.sub.tokens.tf, 2, tf.idf, idf = train.sub.tokens.idf)
-```
-
-```{r}
-# transpose and fix missing values
-test.sub.tokens.tfidf <- t(test.sub.tokens.tfidf)
-test.sub.tokens.tfidf[is.na(test.sub.tokens.tfidf)] <- 0.0
-```
-
-```{r}
-# project the SVD from the train split onto the test split
-test.sub.svd.proj <- projSVD(train.sub.irlba, test.sub.tokens.tfidf)
-```
-
-## Final Data Frames for Predictions
-
-```{r}
-####### Body Token Only #######
-# create data frame of body tokens and labels
-test.body.svd <- cbind(label = test$label,
-                       test.body.svd.proj)
-
-saveRDS(test.body.svd, file = here("results", "large-files", "test-body-svd.rds"))
-
-# create data frame of body tokens and labels
-test.body.svd.length <- cbind(label = test$label,
-                              length = test$bodylength,
-                              test.body.svd.proj)
-
-saveRDS(test.body.svd.length, file = here("results", "large-files", "test-body-svd-length.rds"))
-
-# create data frame of body tokens and labels
-test.body.svd.length.sender <- cbind(label = test$label,
-                                     length = test$bodylength,
-                                     senderdomain = test$senderdomain,
-                                     test.body.svd.proj)
-
-saveRDS(test.body.svd.length.sender, file = here("results", "large-files", "test-body-svd-length-sender.rds"))
-
-####### Subject Token Only #######
-# create data frame of subject tokens and labels
-test.sub.svd <- cbind(label = test$label,
-                      test.sub.svd.proj)
-
-saveRDS(test.sub.svd, file = here("results", "large-files", "test-sub-svd.rds"))
-
-# create data frame of subject tokens and labels
-test.sub.svd.length <- cbind(label = test$label,
-                             length = test$bodylength,
-                             test.sub.svd.proj)
-
-saveRDS(test.sub.svd.length, file = here("results", "large-files", "test-sub-svd-length.rds"))
-
-# create data frame of subject tokens and labels
-test.sub.svd.length.sender <- cbind(label = test$label,
-                                    length = test$bodylength,
-                                    senderdomain = test$senderdomain,
-                                    test.sub.svd.proj)
-
-saveRDS(test.sub.svd.length.sender, file = here("results", "large-files", "test-sub-svd-length-sender.rds"))
-
-####### All Tokens #######
-# create data frame of all tokens and labels
-test.all.svd <- cbind(label = test$label,
-                      test.body.svd.proj,
-                      test.sub.svd.proj)
-
-saveRDS(test.all.svd, file = here("results", "large-files", "test-all-svd.rds"))
-
-# create data frame of all tokens and labels
-test.all.svd.length <- cbind(label = test$label,
-                             length = test$bodylength,
-                             test.body.svd.proj,
-                             test.sub.svd.proj)
-
-saveRDS(test.all.svd.length, file = here("results", "large-files", "test-all-svd-length.rds"))
-
-# create data frame of all tokens and labels
-test.all.svd.length.sender <- cbind(label = test$label,
-                                    length = test$bodylength,
-                                    senderdomain = test$senderdomain,
-                                    test.body.svd.proj,
-                                    test.sub.svd.proj)
-
-saveRDS(test.all.svd.length.sender, file = here("results", "large-files", "test-all-svd-length-sender.rds"))
-```
-
-# Generate Predictions
-
-```{r}
-# predict test data with model
-#rf.body.preds <- predict(rf.body, test.body.svd)
-#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
-#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
-```
-
-The `echo: false` option disables the printing of code (only output is displayed).

### 2024-08-08 code/analysis-code/statistical-analysis.qmd
Final commit

--- 2024-07-18 code/analysis-code/statistical-analysis.qmd
@@
 # project singular value decomp onto test split
-projSVD <- function(trainirlba, testtfidf) {
+projSVD <- function(type, trainirlba, testtfidf) {
   sigma.inverse <- 1 / trainirlba$d
@@
                  trControl = ctrl,
-                 tuneLength = 7)
+                 tuneLength = 7,
+                 verbose = FALSE,
+                 allowParellel = TRUE)
   stopCluster(clus)
@@
                  tuneLength = 7,
-                 tuneGrid = gridSVM)
+                 tuneGrid = gridSVM,
+                 probability = T,
+                 verbose = FALSE,
+                 allowParellel = TRUE)
+  stopCluster(clus)
+  return(model)
+}
+
+# tune an XGBoost model
+customXG <- function(df) {
+  cores <- detectCores()
+  clus <- makeCluster(cores[1] - 1) # determine the optimal number
+                                                   # of cores for parallel processing
+  registerDoParallel(clus)
+  model <- train(label ~ ., 
+                 data = df,
+                 method = "xgbTree",
+                 trControl = ctrl,
+                 tuneGrid = gridXG,
+                 verbose = FALSE,
+                 allowParellel = TRUE)
   stopCluster(clus)
@@
 saveRDS(train.sub.irlba, file = here("results", "large-files", "train-sub-irlba.rds"))
+
+train.sub <- as.data.frame(train.sub.irlba$v)
+train.sub <- train.sub %>%
+  rename_all(~ paste0("Z", seq(1, length(names(train.sub)))))
 ```
@@
 train.sub.svd <- data.frame(label = train$label,
-                            train.sub.irlba$v)
+                            train.sub)
 saveRDS(train.sub.svd, file = here("results", "large-files", "train-sub-svd.rds"))
@@
                                    length = train$bodylength,
-                                   train.sub.irlba$v)
+                                   train.sub)
 saveRDS(train.sub.svd.length, file = here("results", "large-files", "train-sub-svd-length.rds"))
@@
                                           senderdomain = train$senderdomain,
-                                          train.sub.irlba$v)
+                                          train.sub)
 saveRDS(train.sub.svd.length.sender, file = here("results", "large-files", "train-sub-svd-length-sender.rds"))
@@
                             train.body.irlba$v,
-                            train.sub.irlba$v)
+                            train.sub)
 saveRDS(train.all.svd, file = here("results", "large-files", "train-all-svd.rds"))
@@
                                    train.body.irlba$v,
-                                   train.sub.irlba$v)
+                                   train.sub)
 saveRDS(train.all.svd.length, file = here("results", "large-files", "train-all-svd-length.rds"))
@@
                                           train.body.irlba$v,
-                                          train.sub.irlba$v)
+                                          train.sub)
 saveRDS(train.all.svd.length.sender, file = here("results", "large-files", "train-all-svd-length-sender.rds"))
 ```
-
-Having generated structured text data that is also manageable, we can now piece together different data frames consisting of the various feature combinations we would like to assess for modeling.
-Based on the relationships observed in our exploratory data analysis those features are the token data for both email subject and body, the body length of the email, and the sender domain for the email.
 
@@
 
+Finally we'll be using the same SVD process as was used on the training data. We'll project the data using our custom `projSVD()` function defined in the Controls and Functions section above.
+
 ```{r}
 # project the SVD from the train split onto the test split
-test.body.svd.proj <- projSVD(train.body.irlba, test.body.tokens.tfidf)
+test.body.svd.proj <- projSVD("bod", train.body.irlba, test.body.tokens.tfidf)
 ```
@@
 # project the SVD from the train split onto the test split
-test.sub.svd.proj <- projSVD(train.sub.irlba, test.sub.tokens.tfidf)
+test.sub.svd.proj <- projSVD("sub", train.sub.irlba, test.sub.tokens.tfidf)
 ```
@@
 ## Final Data Frames for Predictions
+
+Having now completed the transformations necessary to map the test data into the same space as the training data used for modeling we can proceed with making predictions and assessing model performance.
+The latter of which will be continued in a second `statistical-analysis-continued.qmd` file using the pertinent, saved `.rds` files from this document.
+These files have been excluded from any GitHub commits, because of the generous sizes of each model's data file.
+In order to proceed with the next notebook you'll need to generate the model files on your local machine first.
 
@@
 
+### can't be run in a reasonable timeframe
+### too computationally intensive for a MacBook
 # create data frame of body tokens and labels
-test.body.svd.length.sender <- cbind(label = test$label,
-                                     length = test$bodylength,
-                                     senderdomain = test$senderdomain,
-                                     test.body.svd.proj)
+#test.body.svd.length.sender <- cbind(label = test$label,
+#                                     length = test$bodylength,
+#                                     senderdomain = test$senderdomain,
+#                                     test.body.svd.proj)
 
-saveRDS(test.body.svd.length.sender, file = here("results", "large-files", "test-body-svd-length-sender.rds"))
+#saveRDS(test.body.svd.length.sender, file = here("results", "large-files", "test-body-svd-length-sender.rds"))
 
@@
 
+### can't be run in a reasonable timeframe
+### too computationally intensive for a MacBook
 # create data frame of subject tokens and labels
-test.sub.svd.length.sender <- cbind(label = test$label,
-                                    length = test$bodylength,
-                                    senderdomain = test$senderdomain,
-                                    test.sub.svd.proj)
+#test.sub.svd.length.sender <- cbind(label = test$label,
+#                                    length = test$bodylength,
+#                                    senderdomain = test$senderdomain,
+#                                    test.sub.svd.proj)
 
-saveRDS(test.sub.svd.length.sender, file = here("results", "large-files", "test-sub-svd-length-sender.rds"))
+#saveRDS(test.sub.svd.length.sender, file = here("results", "large-files", "test-sub-svd-length-sender.rds"))
 
@@
 
+### can't be run in a reasonable timeframe
+### too computationally intensive for a MacBook
 # create data frame of all tokens and labels
-test.all.svd.length.sender <- cbind(label = test$label,
-                                    length = test$bodylength,
-                                    senderdomain = test$senderdomain,
-                                    test.body.svd.proj,
-                                    test.sub.svd.proj)
+#test.all.svd.length.sender <- cbind(label = test$label,
+#                                    length = test$bodylength,
+#                                    senderdomain = test$senderdomain,
+#                                    test.body.svd.proj,
+#                                    test.sub.svd.proj)
 
-saveRDS(test.all.svd.length.sender, file = here("results", "large-files", "test-all-svd-length-sender.rds"))
+#saveRDS(test.all.svd.length.sender, file = here("results", "large-files", "test-all-svd-length-sender.rds"))
 ```
-
-# Generate Predictions
-
-```{r}
-# predict test data with model
-#rf.body.preds <- predict(rf.body, test.body.svd)
-#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
-#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
-```
-
-The `echo: false` option disables the printing of code (only output is displayed).

### 2024-08-08 products/manuscript/Manuscript.qmd
Final commit

--- 2024-07-18 products/manuscript/Manuscript.qmd
@@
 author: Sean O'Sullivan
-date: "`r Sys.Date()`"
 format:
@@
 csl: ../../assets/apa.csl
+nocite: |
+  @*
+execute:
+  echo: false
 ---
 
-```{r, echo=FALSE, message=FALSE}
+```{r, message=FALSE}
 # load needed packages. make sure they are installed.
@@
                skimr, # for data exploration and summary
+               pROC,
                ggthemes) # for plot theming
@@
 
-*Write a summary of your project.*
+This study explores the efficacy of advanced machine learning techniques in classifying emails as either spam or legitimate.
+The process begins with tokenization -- converting email text into meaningful tokens.
+Term Frequency-Inverse Document Frequency (TF-IDF) is then applied to transform these tokens into numerical features that reflect their importance across the data set.
+To reduce dimensionality and capture the most significant patterns, Singular Value Decomposition (SVD) is then employed, enhancing the efficiency and performance of the subsequent models.
+Repeated cross-validation is used to ensure robust and reliable model evaluation as three powerful classifiers — Random Forest, Radial Support Vector Machines (SVM), and XGBoost — are trained and compared.
+The performance of each model is assessed using Recall and F1 scores to ensure their ability to effectively distinguish between spam and legitimate emails while erring on the side of maximizing as many true positives as possible.
+The results highlight the potential of combining tokenization, TF-IDF, SVD, and repeated cross-validation in building high-performance, black-box email classification systems.
+Conversely, they demonstrate the difficulty of maintaining human interpretability when data sets are large and dimensionality is high.
 
@@
 
-## General Background Information
+## Description of the Data
 
-*Provide enough background on your topic that others can understand the why and how of your analysis*
-
-## Description of data and data source
-
-The data set(s) that I'd liked to examine are a handful of annotated data sets pertaining to fraudulent (spam) emails.
+The data set(s) that we're examining are a handful of annotated data sets pertaining to fraudulent (spam) emails.
 They come from a handful of sources, the largest of the bunch is from the Conference on Email and Anti-Spam.
@@
 
-## Questions/Hypotheses to be addressed
+## Hypotheses to Be Addressed
 
-I'd like to explore the features that are prevalent in the identification of spam emails.
-Utilizing text analysis, in addition to the other email attributes supplied and engineered, my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for use of automated detection by some popular email providers and large internal IT departments across the corporate landscape.
+The purpose of this project is to explore the features that are prevalent in the identification of spam emails.
+Utilizing text analysis, in addition to the other email attributes supplied and engineered, my end goal is to be able to model a classifier that has a high rate of efficacy in identifying spam emails for use of automated detection by some popular email providers and large internal IT departments across the corporate landscape.
 
-I expect that a number of keywords common in spam emails will be relevant, but it's also likely that potential sender domains, whether the sender and receiver domain are the same, and what day of week or time of day the emails are sent are also highly correlated with intentionally spammy behavior.
+We expect that a number of keywords common in spam emails will be relevant, but it's also likely that potential sender domains, whether the sender and receiver domain are the same, and the length of the email body or subject also highly correlated with intentionally spammy behavior.
 
-While there may be some models better suited for this than others I intend to attempt a handful of classifiers and parameter hyper-tuning in order to compare and determine the best overall predictor of spam emails.
+While there may be some models better suited for this than others, our intent is to attempt a handful of ensemble, rule-based classifiers and parameter hyper-tuning in order to compare and determine the best overall predictor of spam emails.
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
-
-## Data import and cleaning
+## Data Import and Cleaning
 
@@
 
-## Exploratory/Descriptive analysis
+## Exploratory Analysis
 
@@
 However, interestingly the distribution of spam emails is bi-modal -- with a second peak at around the 4,000 character count -- which may indicate that spam vendors are aware of the identification potential of email length and have begun targeting a longer length to avoid notice.
+
 ![](../../results/figures/bodylength-distribution.png){.lightbox}
@@
 In doing so we see some clear indications of which domains are favorable to spam vendors and which are not.
-Notably Gmail fails to make the top 25 domains for spam vendors despite being the far-and-away most common choice for legitimate emails.
+Notably, Gmail fails to make the top 25 domains for spam vendors despite being the far-and-away most common choice for legitimate emails.
 Similarly, Yahoo and Hotmail lead the pack for spam and are not among common domains for legitimate emails.
@@
 
-## Statistical analysis
+## Statistical Analysis
 
-*Explain anything related to your statistical analyses.*
+### Preparing the Data
+
+For the purposes of this project our statistical analysis has been completed on a subset of the overall data to allow for reasonable processing times on consumer grade hardware.
+This extends into some of the choices that were made to exclude especially complex permutations of models whose parameter tuning would take days to complete on such hardware.
+For comments and notes about the specific choices made throughout the analysis process you can reference the `statistical-analysis.qmd` in its entirety.
+The following will be an overview of the data manipulation, feature engineering, dimensionality reduction, and model selection employed within:
+
+We began our analysis by establishing consistent, reusable tuning grids and custom functions for transforming the feature space of our training and testing data.
+In addition, we also established single-line, custom model training functions to improve legibility as we work through various permutations of both features and model types.
+We then proceeded to split our data into training and testing splits of 70% and 30% respectively with the latter withheld for model evaluation and to assess any overfitting issues.
+
+Once we had determined our training and testing splits we began to transform our unstructured text data into a structured feature space.
+This was begun by utilizing the `quanteda` package to tokenize each email body and subject, forcing all characters of each token to be lowercase, removing tokens for common English stopwords, and stemming each token to it's English, semantic root so that they could be aggregated on their shared meanings.
+
+With our unstructured text data now structured as simplified tokens we needed to identify which tokens within the overall text corpus are over represented and therefore provided little analytic value.
+In order to do so, we converted our data to a document-frequency matrix and calculated both the individual term frequencies of each token as well as their inverse document frequencies.
+The end goal of which was to created a Term Frequency - Inverse Document Frequency (TF-IDF) value for each token that penalizes and discounts those tokens that are over represented in our email corpus.
+
+Having standardized the values for each token by utilizing the TF-IDF, we were left with a structured matrix of tokens with incredibly high dimensionality (\>49K tokens) and incredibly high sparsity (\>99.9%).
+While these are common problems when analytically modeling text data, and many machine learning models are robust to data sets where the number of features far exceeds the number of observations, the computational burden for modeling with such high dimensionality is far greater than the typical predictive gains.
+Additionally, with such high dimensionality and sparsity we run the risk of added predictive noise depending on the models employed.
+
+To address these concerns our next step involved utilizing singular value decomposition to condense our features set to only those 300 singular values most capable of explaining the variation within our data.
+This process allowed us to both reduce the dimensionality and sparsity of the data while also simultaneously reducing the amount of noise presented by less significant tokens.
+However, it's worth noting that this comes at the cost of interpretability.
+
+Having completed this process for our training data we proceeded to project our test data into the same, final feature space so that the models can ultimately be used to predict with the new data.
+
+### Modeling
+
+With our final structured data prepared we proceeded with tuning 3 different types of classification models with 6 different combinations of features.
+
+The selected models were:
+
+-   Random Forest
+
+-   Radial Support Vector Machines
+
+-   and XGBoost
+
+These models were selected due to their capacity to handle high dimensional feature sets, their robustness to outlier values, and their ability to make accurate predictions about non-linearly related data.
+
+The selected combinations of features utilized with each model were:
+
+-   Email Body Tokens Only
+
+-   Email Body Tokens with Email Body Length
+
+-   Email Subject Tokens Only
+
+-   Email Subject Tokens with Email Body Length
+
+-   All Email Tokens Only
+
+-   and All Email Tokens with Email Body Length
+
+Additional code is also provided that includes permutations with sender email domains.
+However, due to the high cardinality of that feature and the limited processing power available on consumer grade hardware, the results were not considered in the scope of this project.
 
@@
 
-@tbl-summarytable shows a summary of the data.
+Having completed each model we are now able to assess our model performance and ascertain which model is the most well suited for identification of spam emails.
+Utilizing the withheld test data we created predictions for the test data with each of our tuned models and generate confusion matrices to evaluate the overall performance of each.
 
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
+We'll use this information to help inform which models generalized best when classifying new data and will assess which types of mistakes the models tend to err on the side of -- for our purposes allowing the occasional spam message through is likely better than overzealously blocking legitimate emails from reaching the user's inbox.
 
-## Basic statistical analysis
+In order to determine the model least likely to overzealously block legitimate emails we evaluated the performance of each model on the basis of Recall and F1 score.
+The former ensures we're classifying as many true legitimate emails correctly as possible, while the latter assesses the overall accuracy of our classification while factoring in the weights of any class imbalance.
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
+The following table contains the results of each model:
 
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+```{r}
+cm.df <- readRDS(here("results", "tables", "confusion-matrix-compared.rds"))
+cm.df
+```
 
-## Full analysis
+Here we can see that the XGBoost model with only email body tokens seems to achieve the best balance of both F1 score and Recall. The optimal tuning parameters for this model are listed below:
 
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
+```{r}
+best.params <- readRDS(here("results", "tables", "best-params.rds"))
+best.params
+```
 
-Example @tbl-resulttable2 shows a summary of a linear model fit.
+We can see that when examining the confusion matrix for this particular model permutation only 18 false negatives occurred and equally impressively only 26 false positives occurred.
+This is the exact side of classification error that we would hope to land on with more emails being allowed through than necessarily ought to be, but with few legitimate emails being caught in the model.
+
+```{r}
+cm.best <- readRDS(here("results", "tables", "best-confusion-matrix.rds"))
+cm.best
+```
+
+This high degree of accuracy can also be witnessed when plotting the ROC curve and deriving it's associated AUC value.
+With an exceptionally tight fit with the top-left most corner of the graph, the ROC curve for our best model all but maximizes the AUC value with a value of 99.4%.
+
+```{r}
+rocCurve <- readRDS(here("results", "tables", "roc.rds"))
+par(pty = "s")
+plot.roc(rocCurve, legacy.axes = T, percent = T, xlab = "False Positive Rate", ylab = "True Positive Rate", col = "aquamarine3", lwd = 5, print.auc = T)
+```
+
+Lastly we can begin to peer into the workings of the model by examining the variable importance plot.
+However, we begin to see the pitfalls of SVD on model interpretability.
+While we there are sharp declines in importance after the first two singular values and the second two singular values we're unable to ascertain what actual word choices in the body of the email these singular values are comprised of.
+However, considering the swift decline in importance it may be worth attempting to remodel using this same data without SVD on more powerful hardware so that better intrepretability might be possible.
+
+We do note that none of the highest performing models utilized either email subject tokens or email body length to achieve their scores.
+It's possible that the former only worked to introduce noise to the classification models and the latter was partially encompassed in the analysis of the tokens that comprise the body's length.
+
+![](../../results/figures/final-var-imp.png)
 
@@
 
-# Discussion
+# Conclusions
 
-## Summary and Interpretation
+Many of the limitations of this project were those that are inherent in text analytics and highly dimensional predictive modeling.
+With a very feature rich structured data frame the computational load is quite high.
+In total the modeling sessions utilizing the repeated cross-validation methodology we employed took a little over 3 hours to run through the permutations that contained the above mentioned features.
+There are additional features to explore in future iterations should the necessarily powerful hardware be available such as the inclusion of sender domain in addition to timestamp features like time of day, day of week, and month of year factors.
 
-*Summarize what you did, what you found and what it means.*
+Additionally, the use of singular value decomposition allowed for fairly complex modeling to take place on modest hardware, but it significantly hurt the interpretability of the model.
+While this may work as a black-box solution if rolled out into production, understanding what tokens are most important could be useful to better tune the model in the future by revising and customizing some elements as stopword lists.
 
-## Strengths and Limitations
-
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
-## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+In spite of these limitations, once the models were allowed to run we were able to produce quite sophisticated classification results with few legitimate emails being labeled as spam by our best model within our test set.
 

### 2024-08-08 products/manuscript/Manuscript.qmd
Final commit

@@
   docx:
+    df-print: kable
     toc: false
@@
 cm.df <- readRDS(here("results", "tables", "confusion-matrix-compared.rds"))
-cm.df
+cm.df %>% select(Model, F1, Recall, Precision, `Balanced Accuracy`)
 ```
